---
editor: visual
bibliography: citations.bib
csl: apa-6th-edition.csl

format:
  pdf:
    linkcolor: black
    urlcolor: black
    citecolor: black
    filecolor: black
    toccolor: black
    title-block-style: none
    toc: true
    toc-depth: 4
    number-sections: true
    documentclass: scrreprt
    include-in-header: header.tex
    include-before-body: cover.tex
    preview: false
    dev: cairo_pdf
---

```{r}
#| include: false
library(nutricalc)
library(knitr)
library(dplyr)
library(kableExtra)

```

\clearpage
\pagenumbering{roman}
\setcounter{page}{1}

# Abstract {.unnumbered}

Efficient nutrient-solution formulation is fundamental to controlled horticultural production, yet in practice it is often based on fixed recipes or iterative manual adjustment. This thesis addresses that limitation through the development of a modular open-source computational framework in R for reproducible nutrient-solution design called NutriCalc.

The central objective was to formalize nutrient formulation as a mathematically defined problem that can be solved transparently under agronomic and chemical constraints. To achieve this, the formulation task was expressed as a constrained linear optimization problem and solved using non-negative least squares, ensuring physically meaningful fertilizer quantities. The framework combines stoichiometric conversion, unit transformation, charge-balance-based evaluation and estimation of pH and electrical conductivity, thereby linking nutrient targets to key physicochemical properties of the final solution.

NutriCalc was implemented both as an R package for programmable workflows and as a Shiny-based graphical user interface for interactive use. Overall, the thesis demonstrates that nutrient-solution design can be shifted from a recipe-driven practice to a reproducible, optimization-based and chemically constrained workflow. NutriCalc provides a transparent computational foundation for horticultural research, teaching and applied nutrient management.

\listoffigures

\listoftables

# List of Abbreviations {.unnumbered}

A-BAK

:   Stock solution A \[A-stammlösung / stock tank A\]

B-BAK

:   Stock solution B \[B-stammlösung / stock tank B\]

BK

:   Base Capacity \[Basenkapazität\]

BOKU

:   University of Natural Resources and Life Sciences, Vienna \[Universität für Bodenkultur Wien\]

BWB

:   Berliner Wasserbetriebe

CO~2~(aq)

:   Aqueous carbon dioxide

CAN

:   Calcium ammonium nitrate

CRC

:   CRC Handbook of Chemistry and Physics

CSV

:   Comma-Separated Values

DWC

:   Deep Water Culture

EC

:   Electrical Conductivity

EDTA

:   Ethylenediaminetetraacetic acid

Fe-EDTA

:   Iron ethylenediaminetetraacetate

GUI

:   Graphical User Interface

HU

:   Humboldt-Universität zu Berlin

meq L^-1^

:   Milliequivalents per litre

mmol L^-1^

:   Millimoles per litre

N~A~

:   Avogadro constant

NNLS

:   Non-Negative Least Squares

pH

:   Negative logarithm of the hydrogen ion activity

PHREEQC

:   PH REdox EQuilibrium; Geochemical speciation and reaction-modeling software developed by the U.S. Geological Survey

R

:   The R programming language and software environment for statistical computing

RDW

:   Reuse drainage water

SI

:   International System of Units \[Système international d’unités\]

SK

:   Acid Capacity \[Säurekapazität\]

SSE

:   Sum of Squared Errors

SSRE

:   Sum of Squared Relative Errors

TCE

:   Total Charge Equivalents

WUE

:   Water Use Efficiency

\savefrontmatter
\clearpage
\pagenumbering{arabic}
\setcounter{page}{1}

# Introduction

Precise nutrient-solution formulation is fundamental to controlled horticultural production, particularly in hydroponic and other soilless cultivation systems. In these systems plant nutrition depends on the chemical composition of the supplied solution, so inaccuracies in formulation can directly affect plant performance, experimental validity and resource-use efficiency.

In practice, however, nutrient-solution design remains a complex task. Fertilizer salts often supply multiple nutrients simultaneously, local irrigation water contributes additional ions and target concentrations must be translated across different chemical and commercial representations. As a result, the formulation process is error-prone, when performed manually or with opaque spreadsheet-based workflows, especially when transparency, reproducibility and extensibility are required.

These challenges are particularly relevant in horticultural research, where treatments are often designed to manipulate individual nutrients while keeping other factors as constant as possible. Because ions are chemically coupled through shared salts, electroneutrality and solution chemistry, such designs require computationally explicit methods rather than ad hoc adjustment. Despite this need, no dedicated modular open-source framework in R is currently available for transparent nutrient-solution formulation and analysis in horticultural applications.

This thesis addresses that gap through the development of NutriCalc, an open-source computational framework in R for the formulation, analysis and optimization of nutrient solutions. The work formalizes nutrient formulation as a constrained linear optimization problem solved using non-negative least squares and extends this workflow through stoichiometric conversion, charge-balance-based evaluation, water input and estimation of pH and electrical conductivity. In addition to the computational backend, the framework is implemented as a Shiny-based graphical user interface to support practical use by researchers and horticultural practitioners. The overarching objective of the thesis is to provide a reproducible chemically coherent and extensible framework that shifts nutrient-solution formulation from a recipe-based practice to a transparent computational workflow.

## Gap: lack of transparent, reproducible, flexible formulation tools

A reconstruction of nutrient calculations reported in the literature further illustrates the relevance of transparent conversion pipelines. For example, @Jones.2024 describe a control nutrient solution for CUBES Circle derived from defined molar concentrations of fertilizer salts (5.15 mM Ca(NO₃)₂; 8.79 mM KNO₃; 2.36 mM MgSO₄; 2.94 mM KH₂PO₄), subsequently scaled to a 10 % strength solution. Recalculating the resulting elemental concentrations reported in the paper reveals that the final values depend critically on how stoichiometric relationships, dilution factors and ion attribution are applied. In particular, deviations may arise when dilution factors are inconsistently propagated across elements, when molar masses of salts are used instead of elemental molar masses, or when contributions of shared ions are only partially accounted for. While K, P, and S match the reported control values, the largest discrepancies occur for Ca, Mg, and \ce{NO3-}. Recomputing the resulting concentrations using stoichiometric attribution and elemental molar masses yields:

```{r}
compute_molar_mass("Ca")
```

$$
c(\mathrm{Ca})
= 5.15\,\si{mmol.L^{-1}}
\times 40.08\,\si{mg.mmol^{-1}}
\times 0.1
= 20.64\,\si{mg.L^{-1}}.
$$

```{r}
compute_molar_mass("Mg")
```

$$
c(\mathrm{Mg})
= 2.36\,\si{mmol.L^{-1}}
\times 24.31\,\si{mg.mmol^{-1}}
\times 0.1
= 5.74\,\si{mg.L^{-1}}.
$$

```{r}
compute_molar_mass("NO3")
```

$$
c(\ce{NO3-})
= \bigl(2\times 5.15 + 8.79\bigr)\,\si{mmol.L^{-1}}
\times 62.01\,\si{mg.mmol^{-1}}
\times 0.1
= 118.38\,\si{mg.L^{-1}}.
$$

The reported Ca value ($82.05\ \si{mg.L^{-1}}$) is consistent with using the molar mass of \ce{Ca(NO3)2} (and effectively $5.0\ \si{mM}$ instead of $5.15\ \si{mM}$) rather than elemental Ca:

```{r}
compute_molar_mass("Ca(NO3)2")
```

$$
c(\mathrm{Ca})_{\text{reported}}
= 5.0\,\si{mmol.L^{-1}}
\times 164.09\,\si{mg.mmol^{-1}}
\times 0.1
= 82.05\,\si{mg.L^{-1}}.
$$

The reported Mg value ($57.37\ \si{mg.L^{-1}}$) is consistent with the $0.1$ scaling factor not being applied:

$$
c(\mathrm{Mg})_{\text{reported}}
= 2.36\,\si{mmol.L^{-1}}
\times 24.31\,\si{mg.mmol^{-1}}
= 57.37\,\si{mg.L^{-1}}.
$$

The reported nitrate value ($63.87\ \si{mg.L^{-1}}$) is consistent with counting nitrate only from \ce{Ca(NO3)2} (omitting the \ce{KNO3} contribution):

$$
c(\ce{NO3-})_{\text{reported}}
= \bigl(2\times 5.15\bigr)\,\si{mmol.L^{-1}}
\times 62.01\,\si{mg.mmol^{-1}}
\times 0.1
= 63.87\,\si{mg.L^{-1}}.
$$

While such differences do not necessarily affect the overall conclusions of the study, they demonstrate how small ambiguities in conversion logic can lead to substantial divergence in reported nutrient concentrations. These observations underline that even potentionally well documented experimental studies can benefit from calculation frameworks in which unit conversions, stoichiometric coefficients and scaling factors are explicitly encoded and reproducible. This is particularly relevant for integrated production systems, where nutrient streams are reused across modules and quantitative consistency is essential.

## Background and Motivation

Precise nutrient solution formulation is a cornerstone of modern horticultural production, particularly in hydroponic and soilless cultivation systems. This process involves selecting and combining soluble fertilizer salts to deliver specific concentrations of essential elements in desired chemical forms and ratios [@Jones.2014]. While standard formulations were published since the mid-19th century [@Hoagland.1920; @Hoagland.1938; @Knop.1860; @Shive.1915], practitioners frequently need to tailor them to local water quality, crop-specific requirements or available fertilizers, creating a complex and often error-prone challenge [@Resh.2013; @Sonneveld.2009].

In plant nutrition experiments, treatments are typically designed to vary individual nutrients while holding others constant (ceteris paribus) to investigate the effect of a single nutrient element. However, due to the interdependencies among ions and their shared presence in various fertilizer salts, such isolation is rarely achievable in practice. Researchers must therefore account for the introduction of accompanying ions, which may affect plant responses and experimental validity.

Creating accurate nutrient solutions under such constraints is time-consuming and sometimes mathematically infeasible, underscoring the need for computational tools that enable efficient, accurate, and reproducible formulation under realistic chemical and experimental conditions.

The nutrient formulation problem is inherently mathematical. Many fertilizer salts contribute multiple nutrients (e.g., Ca(NO₃)₂ supplies both calcium and nitrate), leading to an interdependent system. This can be modeled as a system of linear equations:

$A·x = b$

Where:

-   A is the nutrient matrix (each row representing a nutrient, each column a fertilizer salt),

-   x is the vector of unknown fertilizer quantities,

-   b is the vector of target nutrient concentrations.

Due to overlapping contributions and limited degrees of freedom, exact solutions may not exist. Thus, approximate solutions are needed, with the additional constraint that all fertilizer quantities must be non-negative. This motivates the use of non-negative least squares (NNLS) optimization, which minimizes the squared error between the achieved and target nutrient vectors under the constraint x ≥ 0.

Moreover, real-world use requires support for various chemical notations such as conversions between elemental and oxide forms (e.g., P ↔ P₂O₅, S ↔ SO₄²⁻) and integration with commercially available fertilizers labeled in % w/w or expressed in units such as mmol/L or mg/L. Despite the existence of some basic Windows applications and Excel tools, there is currently no open-source, modular, and transparent software solution in R that offers this level of flexibility and chemical accuracy for horticultural applications.

## Objectives and research questions

The main objective of this thesis is to design and implement a computational framework in R that enables the formulation, analysis and optimization of nutrient solutions specifically for controlled-environment horticulture. This framework is intended to support accurate and efficient nutrient planning and to facilitate reproducibility and transparency in both research and practical applications.

More specifically, the thesis aims to develop tools that can construct nutrient matrices from molecular fertilizer formulas. These tools will allow for precise conversion between fertilizer compounds and their elemental forms using stoichiometric calculations. The formulation process will be modelled mathematically as a non-negative linear system and solved using Non-Negative Least Squares optimization methods [@Mullen.2024].

In addition to the computational backend, the framework will include features to visualize the distribution of nutrients across different fertilizer sources. Furthermore, it will offer the ability to approximate the and pH and EC of a nutrient solution in advance, based on the concentrations and molar conductivity of respective ions [@McCleskey.2012].

The tool will also be designed with extensibility in mind, laying the groundwork for the import of commercial fertilizer compositions. In the long term, the framework will be equipped to benchmark nutrient formulations based on either experimental or simulated crop yield data.

Finally, to ensure accessibility and ease of use, the thesis will culminate in the development of a user-friendly graphical interface using the Shiny framework, enabling non-technical users to benefit from the computational functionalities [@Chang.2025].

# State of the Art / Scientific Background

## Principles of nutrient solution formulation and the structure of experimental design in plant nutrition

This chapter addresses horticultural plant nutrition experiments aimed at evaluating nutrient composition and its effects under chemically and statistically controlled conditions. It does not address other types of biological experiments, such as studies on uptake mechanisms, isotope tracing, signaling pathways, or related physiological processes.

Experiments in plant nutrition seek to identify causal relationships between specific nutrients and plant responses such as growth, yield, vitality, and physiological performance. Despite the apparent simplicity of manipulating nutrient concentrations, such experiments are fundamentally constrained by chemical laws that limit causal identifiability. In particular, electroneutrality and the statistical requirement of ceteris paribus jointly restrict how nutrient solutions can be manipulated and how resulting effects may be interpreted. This chapter develops a formal framework, expressed in milliequivalent (meq) notation, that derives classical experimental designs in plant nutrition from first principles and clarifies the interpretational limits of nutrient experimentation.

A nutrient solution may be represented as a vector of its ionic charge equivalents:

```{=latex}
\begin{align*}
\mathbf{S}_{\mathrm{meq}} &=
\bigl(
\mathrm{K^+},\ \mathrm{Ca^{2+}},\ \mathrm{Mg^{2+}},\ \dots
\;\big|\;
\mathrm{NO_3^-},\ \mathrm{Cl^-},\ \mathrm{SO_4^{2-}},\ \dots
\bigr)
\end{align*}
```

where each component is expressed in milliequivalents per liter (meq L$^{-1}$). For any ion $i$,

```{=latex}
\begin{align*}
\mathrm{meq}_i &= |z_i|\, c_i ,
\end{align*}
```

with $z_i$ denoting ionic charge and $c_i$ the molar concentration. Expressing solution composition in meq makes ionic charge balance explicit and avoids ambiguity associated with heterovalent ions.

Not all vectors in this space are chemically admissible. All aqueous nutrient solutions must satisfy electroneutrality, which in meq notation takes the form

```{=latex}
\begin{align*}
\sum \mathrm{meq}_{\text{cat}} - \sum \mathrm{meq}_{\text{an}} &= 0 .
\end{align*}
```

The sums include all charged species present, including $\mathrm{H^+}$ (formally $\mathrm{H_3O^+}$) and $\mathrm{OH^-}$. Electroneutrality implies that no ionic component can be altered independently of ions carrying opposite charge. Any experimental manipulation therefore induces compensatory changes elsewhere in the ionic system. In this chapter, meq values are treated as analytical charge equivalents used for experimental design; activity coefficients, complexation, and ion pairing may alter speciation but do not remove the electroneutrality constraint.

In addition to charge balance, many nutritionally relevant components participate in acid–base equilibria and therefore exist as families of chemical species. For example, phosphate occurs as $\mathrm{H_2PO_4^-}$, $\mathrm{HPO_4^{2-}}$, and $\mathrm{PO_4^{3-}}$, while carbonate occurs as $\mathrm{CO_2}$, $\mathrm{HCO_3^-}$, and $\mathrm{CO_3^{2-}}$. Let $c_i^{\mathrm{tot}}$ denote the total analytical concentration of component $i$. The total charge contribution of this component to the solution can be written as

```{=latex}
\begin{align*}
\mathrm{meq}_i^{\mathrm{tot}}
&= \sum_j |z_{ij}|\, c_{ij} ,
\end{align*}
```

where $c_{ij}$ is the concentration of species $j$ belonging to component $i$. Changes in pH redistribute charge among species with different valencies as shown for phosphate in @fig-phosphate-speciation-tight, thereby altering the effective ionic composition of the solution even when total analytical concentrations are held constant. Although pH can be experimentally controlled using buffers or acid–base additions, such control necessarily introduces additional ionic components and therefore does not decouple pH from nutrient composition in a causal sense.

```{r}
#| echo: false
#| fig-cap: Phosphate speciation (percent) as a function of pH.
#| label: fig-phosphate-speciation-tight
#| fig-height: 4.5
#| fig-width: 7

pKa1 <- 2.15
pKa2 <- 7.20
pKa3 <- 12.35

Ka1 <- 10^(-pKa1)
Ka2 <- 10^(-pKa2)
Ka3 <- 10^(-pKa3)

pH <- seq(0, 14, by = 0.01)
H  <- 10^(-pH)

D <- H^3 + Ka1*H^2 + Ka1*Ka2*H + Ka1*Ka2*Ka3
alpha0 <- H^3 / D
alpha1 <- Ka1*H^2 / D
alpha2 <- Ka1*Ka2*H / D
alpha3 <- Ka1*Ka2*Ka3 / D

pct <- 100 * cbind(alpha0, alpha1, alpha2, alpha3)

op <- par(
  mar = c(4.2, 4.2, 3.0, 1) + 0.1,  # slightly more top margin
  xaxs = "i",
  yaxs = "i"
)

matplot(
  pH, pct, type = "l", lwd = 2, lty = 1, col = "black",
  xlab = "pH", ylab = "Percentage",
  xlim = c(0, 14), ylim = c(0, 100),
  xaxt = "n", yaxt = "n"
)

axis(1, at = 0:14)
axis(2, at = seq(0, 100, by = 10), las = 1)
box(lwd = 1.5)

text(
  x = c(0.8, 5.0, 10.0, 13.5), y = 106,  # moved higher
  labels = c(
    expression(H[3]*PO[4]),
    expression(H[2]*PO[4]^"-"),
    expression(HPO[4]^"2-"),
    expression(PO[4]^"3-")
  ),
  xpd = NA
)

par(op)


```

From a causal-inference perspective, identification of an effect requires isolating variation in a target factor while holding other causally relevant factors fixed or blocked (ceteris paribus). In plant nutrition experiments, strict independent variation of a single ionic species is generally not possible, because electroneutrality and acid–base equilibria couple ionic charge, pH, ionic strength, and osmotic potential. As a result, strict ceteris paribus conditions for individual ions cannot be achieved in bulk nutrient solutions; they can only be approximated within constrained experimental frameworks. Within this framework, classical experimental designs in plant nutrition can be derived as three distinct classes of experiments, defined by the invariants they preserve in meq-space.

Here, $\Delta$ denotes the change induced by the experimental manipulation, i.e. $\Delta x := x_{\text{treatment}} - x_{\text{control}}$ (or, more generally, the difference between two solution states). The first class consists of qualitative, or ion-substitution, experiments. In these experiments, one ionic species is replaced by another while total cation and total anion charge equivalents are kept constant:

```{=latex}
\begin{align*}
\Delta \mathrm{meq}_i + \Delta \mathrm{meq}_j &= 0 .
\end{align*}
```

When the substituted ions have the same charge, the substitution is isovalent (e.g. $\mathrm{K^+} \leftrightarrow \mathrm{Na^+}$ or $\mathrm{Cl^-} \leftrightarrow \mathrm{NO_3^-}$). When the charges differ, the substitution is heterovalent, for example replacing $2$ meq of $\mathrm{K^+}$ (2 mmol) with $2$ meq of $\mathrm{Ca^{2+}}$ (1 mmol). Qualitative substitution experiments provide the highest degree of causal specificity achievable within whole-solution manipulations at fixed charge totals, although their interpretation remains conditional on buffering and counter-ion choice.

The second class comprises quantitative, or salt addition (or removal), experiments. Here, the total amount of dissolved salt is changed, which necessarily alters both total cation and total anion charge equivalents by the same non-zero amount:

```{=latex}
\begin{align*}
\Delta \sum \mathrm{meq}_{\text{cat}}
&= \Delta \sum \mathrm{meq}_{\text{an}}
\neq 0 .
\end{align*}
```

Electroneutrality is preserved automatically, but ionic strength, electrical conductivity, and osmotic potential change and pH may shift depending on buffering. Because at least two ionic components vary simultaneously, plant responses in such experiments reflect composite effects of nutrient ions, counter-ions and general salinity.

The third class treats quantity as an emergent quality through solution-strength series. In this case, all charge equivalents in the solution are scaled by a constant factor $k$:

```{=latex}
\begin{align*}
\mathbf{S}_{\mathrm{meq},k}
&= k\, \mathbf{S}_{\mathrm{meq},1} .
\end{align*}
```

This transformation preserves ion ratios expressed in meq while increasing or decreasing total charge density, ionic strength and osmotic potential. Although composition remains constant at the level of charge ratios, pH and ionic speciation may shift because acid–base equilibria are nonlinear. As a result, solution strength functions as an environmental condition rather than a simple quantitative variable.

Across these three experiment types, causal specificity for individual ion identity typically decreases as the number of inseparable or uncontrolled variables increases. Qualitative substitution experiments offer the highest degree of causal specificity, whereas quantitative salt manipulations and solution-strength series progressively trade causal resolution for ecological or agronomic relevance. These experiment classes are therefore best understood as distinct projections of a chemically constrained ionic system, most transparently described in milliequivalent space.

\begin{landscape}
\begingroup
\fontsize{11}{13}\selectfont
\setlength{\tabcolsep}{4pt}
\renewcommand{\arraystretch}{1.1}

\begin{tabular}{
>{\raggedright\arraybackslash}p{2.8cm}
>{\raggedright\arraybackslash}p{3.2cm}
>{\raggedright\arraybackslash}p{4.2cm}
>{\raggedright\arraybackslash}p{3.2cm}
>{\raggedright\arraybackslash}p{4.0cm}
>{\raggedright\arraybackslash}p{2.7cm}
>{\raggedright\arraybackslash}p{3.2cm}
}
\toprule
Experiment class & Core manipulation & Charge balance condition & Held constant & Necessarily changes & Example & Interpretation \\
\midrule
I.a Qualitative (isovalent) &
Equal-meq substitution &
$\begin{aligned}
\Delta \sum \mathrm{meq}_{cat} &= 0 \\
\Delta \sum \mathrm{meq}_{an}  &= 0
\end{aligned}$ &
Total cation \& anion meq &
Ion identity &
$\mathrm{K^+}\!\leftrightarrow\!\mathrm{Na^+}$ &
Conditional ion identity effect \\
\addlinespace[3pt]
I.b Qualitative (heterovalent) &
Equal-meq substitution &
$\Delta \mathrm{meq}_i+\Delta \mathrm{meq}_j=0$ &
Charge equivalents &
Molar conc., ionic strength &
$2$ mmol $\mathrm{K^+}\!\leftrightarrow\!1$ mmol $\mathrm{Ca^{2+}}$ &
Identity + secondary effects \\
\addlinespace[3pt]
II Quantitative &
Salt addition or removal &
$\begin{aligned}
\Delta \sum \mathrm{meq}_{cat} &= \\ \Delta \sum \mathrm{meq}_{an} 
&\neq 0
\end{aligned}$ &
Electroneutrality &
Ions, EC, osmotic potential &
$\uparrow\,\mathrm{KCl}$ &
Composite salt effect \\
\addlinespace[3pt]
III Quantitative-as-qualitative &
Solution scaling &
$\mathbf{S}_{\mathrm{meq},k}=k\,\mathbf{S}_{\mathrm{meq},1}$ &
meq ratios &
Total charge, speciation &
¼×, ½×, 1× &
Environmental strength effect \\
\bottomrule
\end{tabular}

\endgroup
\end{landscape}

From a plant nutrition perspective, comparisons between nutrient solutions of different provenance are only informative to the extent that differences in ionic composition are maintained or eliminated in a controlled manner. If two solutions of different origin, such as wastewater-derived (e.g. animal or human urine) and mineral solutions, are formulated to the same ionic composition, then the specifically nutritional dimension of the comparison is largely removed, reducing the question to a Ship of Theseus problem of identity. Under such conditions, any remaining differences are more appropriately interpreted in terms of economic factors associated with source, production, transport, cost and sustainability rather than nutrient composition itself.

## Stoichiometry and salt–ion relationships

In hydroponics, nutrients are supplied by dissolving fertilizer salts in water to prepare a nutrient solution [@Resh.2013]. The formulation of such solutions is complicated by the declaration conventions used for commercial fertilizers. Nutrient contents are commonly expressed in oxide form rather than as elemental concentrations or ionic species. Thus, potassium is typically declared as K₂O rather than as K or ionic K⁺, while phosphorus is conventionally declared as P₂O₅ rather than as P or as the pH-dependent phosphate species H₂PO₄⁻, HPO₄²⁻, or PO₄³⁻. Although these conventions are historically established, they introduce an unnecessary level of abstraction and increase the susceptibility of nutrient calculations to error.

A more serious limitation is that fertilizer labels do not necessarily disclose all contained elements. As a result, attempts to recalculate complete fertilizer compositions in ionic form often fail to account for 100 % of the product mass. One reason is that elements such as chloride may be present without mandatory declaration. Under Regulation (EC) No 2003/2003, only specific nutrient categories are subject to declaration requirements, namely the primary nutrients nitrogen, phosphorus, and potassium; the secondary nutrients calcium, magnesium, sodium, and sulfur; and the micronutrients boron, cobalt, copper, iron, manganese, molybdenum, and zinc. In addition, secondary nutrients must only be declared above specified threshold concentrations. For sodium, for example, declaration is required only above 3 % Na₂O or 2.2 % Na.

For this reason, the present work was restricted to single salts or mixtures of single salts with chemically defined composition. Such compounds can be represented directly by their molecular formulas, so that molar mass and stoichiometric composition are fully determined by the sum of the constituent elements and their stoichiometric coefficients. This ensures complete compositional transparency and avoids the uncertainty associated with incomplete commercial fertilizer declarations.

Many fertilizers used in greenhouse production are in fact single salts. A representative example is Epsom salt, MgSO₄·7H₂O, marketed by Yara as YaraTera KRISTA MgS and declared as containing 16 % MgO and 32.5 % SO₃ [@YARA.2026]. Chemically, pure MgSO₄·7H₂O contains 9.86 % Mg and 13.01 % S, corresponding to 16.35 % MgO and 32.48 % SO₃, or 38.98 % SO₄. Thus, 51.16 % of the compound mass consists of water of crystallization. Another example is calcium nitrate, which may occur as anhydrous Ca(NO₃)₂, as the hydrate Ca(NO₃)₂·4H₂O, or as the double salt 5Ca(NO₃)₂·NH₄NO₃·10H₂O. These distinctions are stoichiometrically relevant because each form differs in molar mass and nutrient contribution, even where the commercial designation is similar.

## Existing tools

A classical approach to nutrient-solution calculation is the universal algorithm proposed by @Sonneveld.1999. The method follows an iterative balancing procedure in which fertilizer salts are added stepwise to meet nutrient targets, while the residual nutrient requirements are recalculated after each addition because each salt simultaneously contributes several ions. In this way, the final formulation is obtained through successive correction of remaining deficits under the constraints imposed by salt composition, target EC and raw-water composition.

Nutrient Solution Calculator is an Excel™-based tool developed by Luca Incrocci and supported by the EUPHOROS project [@Pardossi.2011]. It was previously available for download via Wageningen University and was still accessible in 2024, but was no longer publicly available there in 2026 [@Incrocci.2007]. A Spanish translation remains available online through the University of Almeria [@Incrocci.2020]. The spreadsheet is designed to calculate the composition and cost of fertigation stock solutions (A+B) from three principal inputs: the ionic composition of the irrigation water, the crop nutrient recipe, including target pH, EC, and nutrient concentrations and the technical characteristics of the fertigation system, particularly stock tank volume and dilution ratio. Furthermore, it evaluates the potential risk of salt precipitation in concentrated stock solutions. The spreadsheet is password-protected.

Another tool is HydroBuddy, a free and open-source software application developed by Daniel Fernández Pinto for the formulation of hydroponic nutrient solutions [@FernandezPinto.2022]. The program calculates the required quantities of fertilizers to achieve specified target nutrient concentrations. It includes a database of commonly used fertilizers whose compositions can be edited or extended by the user and allows the formulation of both direct nutrient solutions and concentrated stock solutions. The software determines the combination of fertilizer salts that best matches the desired nutrient composition through the solution of a system of linear equations and also provides an empirical estimation of the EC.

At present, no dedicated open-source R library is available for transparent, modular nutrient-solution calculation in horticultural applications.

# Material and Methods

The development of the computational framework followed a structured development and validation pipeline. The R package was developed using the devtools package [@Wickham.2025].The first step involved designing the underlying data structures. A nutrient matrix was established in which the rows represented nutrients (such as N, P, K, and Ca) and the columns represented fertilizer salts. For readability, the input matrix was stored in transposed form and converted into the required orientation for calculation. Each fertilizer was parsed on the basis of its molecular formula into its elemental components using a database of molar masses and established stoichiometric rules.

Building on this foundation, an R package was developed. The package includes core functions for parsing chemical formulas and computing molar masses, as well as for converting between different chemical representations such as oxides and elemental forms (e.g., P₂O₅ to P or SO₄ to S). It also includes tools for transforming concentration units (e.g., from mmol L⁻¹ to mg L⁻¹) using accurate molar-mass calculations. In addition, bidirectional transformations between compound-based and elemental quantities were integrated on the basis of stoichiometric relationships.

The central component of the package was the implementation of the nutrient-solution formulation algorithm. This algorithm used nnls::nnls() to solve the system of equations under non-negativity constraints and to quantify deviations between the target and calculated solution in terms of absolute and relative error. The framework also allows users to assign weights to individual nutrients, thereby enabling the exploration of different formulation priorities and trade-offs.

The framework was validated through a series of use cases. These included reproducing historical and crop-specific nutrient solutions, comparing the computed results with values reported in the literature, and compiling crop- and author-specific nutrient solutions into a base library. The distribution of anions and cations across different solutions was visualized, for example, by ternary plots for cations (K⁺, Ca²⁺, Mg²⁺) and anions (NO₃⁻, SO₄²⁻, PO₄³⁻), in order to classify nutrient combinations systematically, as proposed by @Steiner.1961.
## Software Environment and Data Sources

### R version, packages

The package `nutricalc` was written in R 4.5.1 [@RCoreTeam.2025] using Posit RStudio 2025.09.2+418 "Cucumberleaf Sunflower" [@Positteam.2025]. Following packages were used: `stringr` [@Wickham.2025], `nnls` [@Mullen.2024], `shiny` [@Chang.2025], `bslib` [@Sievert.2025], `Ternary` [@Smith.2025], `xml2` [@Wickham.2026]. For the development in R `devtools` [@Wickham.2025b] was used.

Code of the package is provided on GitHub under <https://github.com/diolsa/nutricalc> and the shiny-App is deployed under: <https://nutricalc.shinyapps.io/nutricalc-app/>

Code development was supported by AI-assisted LLM for debugging and optimization to ensure code quality and computational reliability with OpenAI CODEX and ChatGPT 5.2.

### Built-in recipe datasets

@Hewitt.1966 compiled a large proportion of the nutrient-solution formulations used during the first half of the twentieth century. For the present work, selected standard recipes from this compilation were considered, particularly those that remain relevant in contemporary research and horticultural practice. In addition, 38 crop-specific horticultural recipes from @Sonneveld.1994 were included. These recipes were assigned to the three groups `Olericulture`, `Floriculture`, and `Fruticulture`. In the original source, @Sonneveld.1994 also classified the formulations into the categories `A`, `B`, and `C`, reflecting their practical and experimental suitability for the respective crop. The resulting set of formulations is shown in @tbl-crop-recipes-long-landscape_units.

```{r}
#| label: tbl-crop-recipes-long-landscape_units
#| echo: false
#| message: false
#| warning: false
#| results: asis

library(dplyr)
library(kableExtra)

# ---- ensure objects exist ----
if (!exists("recipes") || !exists("CANONICAL_NUTRIENTS")) {
  if (requireNamespace("nutricalc", quietly = TRUE)) {
    library(nutricalc)
  } else {
    source("R/recipes.R")
  }
}

nutrients <- CANONICAL_NUTRIENTS

fmt <- function(x) formatC(x, format = "f", digits = 6, drop0trailing = TRUE)

clean_name <- function(x) {
  x <- gsub("^[^A-Za-z]+\\s*", "", x)
  x <- gsub("\\(Reuse Drainage Water\\)", "(RDW)", x)
  x
}

cat_order <- c("Olericulture", "Floriculture", "Fruticulture")
exclude_recipes <- c("lettuce_bht")

micro_nutrients <- c("Fe","Mn","Zn","B","Cu","Mo","Si")

# ---- build data frame ----
rows <- list()

for (cat_name in cat_order) {
  recs <- recipes[vapply(recipes, function(r) identical(r$category, cat_name), logical(1))]
  recs <- recs[!names(recs) %in% exclude_recipes]

  for (r in recs) {
    t <- r$targets[nutrients]

    vals_num <- as.numeric(t)
    names(vals_num) <- names(t)

    micros_present <- intersect(names(vals_num), micro_nutrients)
    vals_num[micros_present] <- vals_num[micros_present] * 1000

    vals <- setNames(fmt(vals_num), names(vals_num))

    rows[[length(rows) + 1]] <- c(
      Category = cat_name,
      Recipe   = clean_name(r$name),
      as.list(vals)
    )
  }
}

df <- bind_rows(lapply(rows, as.data.frame.list, stringsAsFactors = FALSE))

num_mat <- suppressWarnings(sapply(df[nutrients], as.numeric))
keep_cols <- colSums(!is.na(num_mat) & num_mat != 0) > 0
df <- df[, c("Category","Recipe", nutrients[keep_cols]), drop = FALSE]

# ---- unit header spans ----
macro_nutrients <- c("NO3_N","NH4_N","P","K","Ca","Mg","S")
micro_nutrients <- c("Fe","Mn","Zn","B","Cu","Mo","Si")

nut_cols   <- setdiff(names(df), c("Category","Recipe"))
macro_cols <- intersect(macro_nutrients, nut_cols)
micro_cols <- intersect(micro_nutrients, nut_cols)
other_cols <- setdiff(nut_cols, c(macro_cols, micro_cols))

lab_macro <- "mmol~L⁻¹"
lab_micro <- "µmol~L⁻¹"

units_header <- c(" " = 2)
if (length(macro_cols) > 0) units_header <- c(units_header, setNames(length(macro_cols), lab_macro))
if (length(micro_cols) > 0) units_header <- c(units_header, setNames(length(micro_cols), lab_micro))
if (length(other_cols) > 0) units_header <- c(units_header, setNames(length(other_cols), " "))

# ---- spacing tweaks ----
cat("\\setlength{\\LTpre}{-10pt}\n")
cat("\\setlength{\\LTpost}{0pt}\n")
cat("\\renewcommand{\\arraystretch}{0.85}\n")

pretty_names <- names(df)
pretty_names[pretty_names == "NO3_N"] <- "NO₃–N"
pretty_names[pretty_names == "NH4_N"] <- "NH₄–N"

tbl <- knitr::kable(
  df,
  format = "latex",
  booktabs = TRUE,
  longtable = TRUE,
  col.names = pretty_names,
  align = c("l","l",rep("c", ncol(df)-2)),
  caption = "Target compositions of crop-specific nutrient solutions (Olericulture, Floriculture, Fruticulture) from @Sonneveld.1994",
  escape = TRUE   # ← IMPORTANT (let kable escape underscores)
) |>
  add_header_above(units_header, escape = FALSE, line = TRUE) |>
  kable_styling(
    latex_options = c("repeat_header"),
    font_size = 7
  ) |>
  collapse_rows(
    columns = 1,
    latex_hline = "major",
    valign = "top"
  ) |>
  landscape()

print(tbl)

cat("\\renewcommand{\\arraystretch}{1}\n")
```

## Data Structures and Chemical Representation

### Elementary Entities

After the 2019 revision of the International System of Units, the base units are defined by fixed numerical values of fundamental constants [@BIPM.2019]. The mole is the SI unit for the amount of a substance; its symbol is mol. The defining constant is the Avogadro constant, N~A~ Thus:

\begin{equation} 1\,\mathrm{mol} = \frac{6.02214076 \times 10^{23}}{N_\mathrm{A}} \end{equation}

An elementary entity may be an atom, a molecule, or any other particle. In this work, the elementary entities are ions as listed in Table \ref{tab:element-molar-mass}. The molar masses are taken from the CRC Handbook of Chemistry and Physics [@Haynes.2017] and rounded to three decimal places. Not all elements reported are essential in plant nutrition. The list include also non-essential or even toxic elements, but elements that part of the plant ionome and potentially present in fertilizers or water even though not all elements reported here where integrated later.

```{r}
#| echo: false
#| results: asis

element_names <- c(
  H  = "Hydrogen",
  C  = "Carbon",
  N  = "Nitrogen",
  O  = "Oxygen",
  P  = "Phosphorus",
  S  = "Sulfur",
  K  = "Potassium",
  Ca = "Calcium",
  Mg = "Magnesium",
  Na = "Sodium",
  Cl = "Chlorine",
  Fe = "Iron",
  Mn = "Manganese",
  Cu = "Copper",
  Zn = "Zinc",
  Mo = "Molybdenum",
  B  = "Boron",
  Al = "Aluminum",
  Co = "Cobalt",
  Se = "Selenium",
  Si = "Silicon",
  Ni = "Nickel"
)

tab_df <- element_molar_mass_df |>
  dplyr::mutate(
    `Element name` = unname(element_names[Element])
  ) |>
  dplyr::rename(
    `Symbol` = Element,
    `Molar Mass [$\\mathrm{g\\,mol^{-1}}$]` = MolarMass_g_per_mol
  ) |>
  dplyr::select(Symbol, `Element name`, `Molar Mass [$\\mathrm{g\\,mol^{-1}}$]`)

tab <- tab_df |>
  knitr::kable(format = "latex", booktabs = TRUE, escape = FALSE, longtable = FALSE)

cat(paste0(
  "\\par\\noindent\n",
  "\\begingroup\n",
  "\\footnotesize\n",
  "\\noindent\\makebox[\\linewidth][c]{%\n",
  "\\begin{minipage}{\\minof{\\linewidth}{\\widthof{", tab, "}}}\n",
  "\\captionsetup{format=plain,labelsep=colon,labelfont=bf,justification=raggedright,singlelinecheck=false,skip=4pt}\n",
  "\\captionof{table}{Molar masses of elementary entities used for nutrient formulation.}\\label{tab:element-molar-mass}\n",
  tab, "\n",
  "\\end{minipage}}\n",
  "\\endgroup\n",
  "\\par\n"
))
```

### Nutrient set

In Table \ref{tab:salt-molar-mass} all implemented 40 fertilizers accesible by the solver are listed with formula, name and respective molar mass. The molar mass is calculated from the chemical formula using the molar masses of the constituent elements and their stoichiometric coefficients reported in \ref{tab:element-molar-mass}.

```{r}
#| echo: false
#| results: asis

pretty_formula_tex <- function(x) {
  x <- as.character(x)

  # mhchem doesn't like the Unicode middle dot (·) → use mhchem hydrate operator "*"
  x <- gsub("\u00b7", "*", x, fixed = TRUE)

  # (optional) normalize any strange unicode minus just in case
  x <- gsub("\u2212", "-", x, fixed = TRUE)

  paste0("\\ce{", x, "}")
}


salt_tbl <- salt_info |>
  mutate(
    Formula = pretty_formula_tex(salt),
    `Molar mass [g mol\\textsuperscript{-1}]` = sapply(salt, compute_molar_mass)
  ) |>
  select(
    Formula,
    Name = category,
    `Molar mass [g mol\\textsuperscript{-1}]`
  ) |>
  arrange(Name)

tab <- salt_tbl |>
  kable(
    format = "latex",
    booktabs = TRUE,
    escape = FALSE,   # IMPORTANT: allow \ce{...}
    digits = 3
  )

cat(paste0(
  "\\par\\noindent\n",
  "\\begingroup\n",
  "\\footnotesize\n",
  "\\noindent\\makebox[\\linewidth][c]{%\n",
  "\\begin{minipage}{\\minof{\\linewidth}{\\widthof{", tab, "}}}\n",
  "\\captionsetup{format=plain,labelsep=colon,labelfont=bf,justification=raggedright,singlelinecheck=false,skip=4pt}\n",
  "\\captionof{table}{Fertilizer compounds and their molar masses.}\\label{tab:salt-molar-mass}\n",
  tab, "\n",
  "\\end{minipage}}\n",
  "\\endgroup\n",
  "\\par\n"
))

```

### Compound-by-nutrient matrix

Table \ref{tab:nutrient-matrix-no-chelate} shows the compound-by-nutrient matrix, listing the amount of each nutrient in mmol provided per mmol of compound for all 40 fertilizers. The table is organised with the chemical formula of each compound in the first column, followed by a column indicating the solver stage (1 for salts, 2 for acids and bases) and then columns for each nutrient. Chelates are also part of the full matrix, but are excluded from the table shown here for better readability. Each compound is additionally linked to metadata such as compound name, category and acid/base status.

```{r}
#| echo: false
#| results: asis

pretty_formula_tex <- function(x) {
  x <- as.character(x)
  x <- gsub("\u00b7", "*", x, fixed = TRUE)  # mhchem hydrate operator
  x <- gsub("\u2212", "-", x, fixed = TRUE)
  paste0("\\ce{", x, "}")
}

cols_drop <- c("EDTA", "DTPA", "EDDHA", "HBED")

nm_no_chelate <- nutrient_matrix[, !(colnames(nutrient_matrix) %in% cols_drop), drop = FALSE]
nm_tbl <- as.data.frame(nm_no_chelate)

# ---- solver stage (1 salts, 2 acids/bases)
stage_vec <- ifelse(
  salt_info$is_acid_base[match(rownames(nm_no_chelate), salt_info$salt)],
  2L, 1L
)

# ---- build table: Formula + Stage + nutrients
nm_tbl <- cbind(
  Formula = pretty_formula_tex(rownames(nm_no_chelate)),
  Solver   = stage_vec,
  nm_tbl
)


# ---- escape remaining underscores
colnames(nm_tbl) <- gsub("_", "\\_", colnames(nm_tbl), fixed = TRUE)

tab <- knitr::kable(
  nm_tbl,
  format   = "latex",
  booktabs = TRUE,
  escape   = FALSE,
  digits   = 0,
  row.names = FALSE   
)


cat(paste0(
  "\\par\\noindent\n",
  "\\begingroup\n",
  "\\scriptsize\n",
  "\\setlength{\\tabcolsep}{3pt}\n",
  "\\renewcommand{\\arraystretch}{0.9}\n",
  "\\noindent\\makebox[\\linewidth][c]{%\n",
  "\\begin{minipage}{\\linewidth}\n",
  "\\captionsetup{format=plain,labelsep=colon,labelfont=bf,justification=raggedright,singlelinecheck=false,skip=4pt}\n",
  "\\captionof{table}{Compound--nutrient stoichiometric matrix (mmol nutrient per mmol compound). Stage 1 = salts, Stage 2 = acids/bases.}\\label{tab:nutrient-matrix-no-chelate}\n",
  "\\resizebox{\\linewidth}{!}{%\n",
  tab, "\n",
  "}\n",
  "\\end{minipage}}\n",
  "\\endgroup\n",
  "\\par\n"
))


```

## Algorithmic Methods

### Formula parsing and molar mass calculation

Describes rules and algorithms for parsing chemical formulas and calculating molar masses, including validation and error handling.

```{r}
library(nutricalc)
compute_molar_mass("Ca(NO3)2·4H2O")
```

\newpage

### Stoichiometric conversions

Numerically, the official SI unit for the substance concentration is mol m⁻³; however, this unit is equivalent to mmol l⁻¹. Therefore, the following relationships for conversion between mmol l⁻¹ and mg l⁻¹ were applied, following Appelo and Postma [@Appelo.2005]:

```{=latex}

\begin{align*}
\mathrm{mg\ l^{-1}}   &= \mathrm{mmol\ l^{-1}} \times \mathrm{M} \\
\mathrm{mmol\ l^{-1}} &= \frac{\mathrm{mg\ l^{-1}}}{\mathrm{M}}
\end{align*}

\noindent where \begin{align*}
\mathrm{M} &\text{ is the molar mass of the corresponding element } [\mathrm{g\,mol^{-1}}].
\end{align*}
```

Within the library, conversion between mmol l^-1^ and mg l^-1^ as elements can be accomplished with the convert_units() function. When the input is provided as a named numeric vector, the names are interpreted as nutrient identifiers (e.g., `Ca`, `P`, `NO3_N`) and are internally mapped to their corresponding element symbols to select the appropriate molar mass (from element_molar_mass_df) for conversion. For example, converting a complete set of recipe values from mg l⁻¹ to mmol l⁻¹:

```{r}
convert_units(c(Ca = 211, P = 46), to = "mmol/L")
```

The conversion can also be applied to a single scalar concentration. In this case, the element must be specified explicitly, since a scalar input carries no nutrient name from which the element can be inferred. The element may be supplied either as a named argument or positionally in the order `(x, element, to)`:

```{r}
convert_units(x = 15, element = "N", to = "mg/L")
```

For reporting and verification a recipe can be summarized with the recipe_summary() function:

```{r}
recipe_summary(c(
    NO3_N = 15,
    NH4_N = 1,
    P     = 1.5,
    K     = 6,
    Ca    = 4,
    Mg    = 2,
    S     = 2),
  unit = "mmol/L")
```

When concentrations are declared in mg l⁻¹, masses are often reported in oxidized or compound forms (e.g. SO₄ or MgO). To convert between such compound-based representations and their elemental equivalents, the converte() function can be used:

```{=latex}
\begin{align*}
\mathrm{x_{to}} &= \mathrm{x_{from}} \times
\frac{\mathrm{M_{to}}}{\mathrm{M_{from}}} \times
\frac{\boldsymbol{\nu_{from,E}}}{\boldsymbol{\nu_{to,E}}}
\end{align*}

\noindent
where
\begin{align*}
\mathrm{x_{from}} &\text{ is the quantity expressed as the source formula (input value),} \\
\mathrm{x_{to}}   &\text{ is the equivalent quantity expressed as the target formula (output value),} \\
\mathrm{M_{from}} &\text{ is the molar mass of the source formula } [\mathrm{g\,mol^{-1}}], \\
\mathrm{M_{to}}   &\text{ is the molar mass of the target formula } [\mathrm{g\,mol^{-1}}], \\
\boldsymbol{\nu_{from,E}} &\text{ is the number of atoms of the anchor element } E \text{ in the source formula } [\text{dimensionless}], \\
\boldsymbol{\nu_{to,E}}   &\text{ is the number of atoms of the anchor element } E \text{ in the target formula } [\text{dimensionless}].
\end{align*}
```

```{r}
converte("SO4", "S")
```

It is also possible to specify an explicit quantity to be converted, for example when converting a given mass concentration

```{r}
converte("N", "NO3", 210)
```

If multiple elements are shared between the source and target formulas, the anchor element E must be specified to avoid ambiguity.

```{r}
converte("(NH4)2HPO4", "NH4H2PO4", element = "P")  # P as the anchor
```

### NNLS optimization with weighting

NNLS is a numerical optimization algorithm for solving linear least-squares problems under non-negativity constraints. It was written in 1973 in FORTRAN 77 by Lawson and Hanson at the Jet Propulsion Laboratory and published in their book *Solving Least Squares Problems* [@Lawson.1974; @Lawson.1995]. In R, the Lawson–Hanson algorithm is implemented by the package `nnls` [@Mullen.2024].

For nutrient solution formulation, the objective is to determine the amounts of salts that fit a specified target nutrient composition as closely as possible. Each salt contributes fixed stoichiometric amounts of ions, so the relationship between the amount of each salt and the resulting nutrient concentration in solution is linear. Because salts can only be added and not removed, their amounts are constrained to be non-negative. Nutrient solution design can therefore be formulated as a linear mixture fitting problem subject to non-negativity constraints, which can be solved using the NNLS algorithm.

Because many fertilizer salts supply multiple nutrients simultaneously exact matching of all nutrient targets is rarely possible. The optimization therefore identifies the best feasible compromise among competing nutrient constraints.

Assume that $n$ fertilizer salts are available and $m$ nutrients are considered.

In `NutriCalc`, fertilizer stoichiometry is stored in a matrix $$
S \in \mathbb{R}^{n \times m},
$$ where rows correspond to salts ($j = 1,\ldots,n$) and columns correspond to nutrients ($i = 1,\ldots,m$). The entry $S_{j i}$ gives the amount of nutrient $i$ supplied by one unit of salt $j$ (mmol nutrient per mmol salt).

Let

-   $x \in \mathbb{R}_{\ge 0}^n$ denote the vector of salt amounts (mmol $L^{-1}$),
-   $b \in \mathbb{R}^m$ denote the vector of target nutrient concentrations (mmol $L^{-1}$).

For fitting, the solver uses the transposed stoichiometry matrix $$
A = S^\top \in \mathbb{R}^{m \times n},
$$ so that rows correspond to nutrients and columns correspond to salts. The system therefore consists of $m$ nutrient equations and $n$ non-negative decision variables. Predicted (achieved) nutrient concentrations are $$
\hat b = A x .
$$

Thus, each row of $A$ represents the linear contribution of all salts to one nutrient, and each column represents the nutrient composition of one salt. The optimization therefore selects salt amounts so that the combined nutrient profile best approximates the desired composition.

------------------------------------------------------------------------

### Target scaling and importance weighting

`NutriCalc` does not solve the unweighted least-squares problem $\min_{x \ge 0}\lVert Ax - b\rVert_2$. Instead, residuals are rescaled relative to target magnitude and weighted according to user-defined nutrient importance.

Conceptually, the solver minimizes the total weighted squared deviation between achieved and target nutrient concentrations. For nutrients with nonzero targets, deviations are measured relative to the target (i.e. proportional error). For nutrients with zero targets, deviations are measured in absolute concentration units.

This relative scaling is necessary because nutrient concentrations often differ by several orders of magnitude (for example, macronutrients versus micronutrients). Without scaling, nutrients present at large absolute concentrations would dominate the optimization and prevent meaningful fitting of low-concentration nutrients.

Scaling and importance weighting operate on residuals rather than on concentrations themselves. Optimization decisions therefore depend on deviations from targets rather than absolute concentration levels.

------------------------------------------------------------------------

#### Importance weighting of nutrient targets

In practical nutrient formulation, not all targets must be matched with the same precision. Some nutrients are critical for plant performance and must be maintained close to their target concentration, whereas others can vary within a broader range without adverse effects. The optimization therefore allows nutrient-specific importance levels that determine how strongly deviations from individual targets are penalized.

User-defined importance values $I_i$ express the desired strictness of matching each nutrient. Higher importance means that deviations from the target are penalized more strongly, forcing the solver to prioritize matching that nutrient even if this leads to larger deviations in less important nutrients.

Importance values are restricted to the interval $[-2,2]$ and converted into positive numerical weights using an exponential mapping: $$
\tilde I_i = \min\!\bigl(2,\max(-2,I_i)\bigr),
$$ $$
w_i = 0.1 \cdot 1000^{\tilde I_i}.
$$

Each unit increase in importance multiplies the penalty on deviations by a factor of 1000. This creates clear separation between importance levels and ensures that highly important nutrients strongly influence the optimization outcome. This mapping produces weights spanning approximately $10^{-7}$ to $10^{5}$.

The weights are incorporated into the least-squares objective using the diagonal matrix $$
W = \mathrm{diag}\!\left(\sqrt{w_1},\ldots,\sqrt{w_m}\right),
$$ which scales the residuals before minimization. The optimization therefore minimizes a weighted sum of squared deviations, where each nutrient contributes proportionally to its assigned importance.

------------------------------------------------------------------------

#### Target-based scaling of residuals

Nutrient targets differ substantially in magnitude, with macronutrients typically present at concentrations several orders of magnitude larger than micronutrients. If deviations were measured only in absolute units (mmol L$^{-1}$), nutrients with large target concentrations would dominate the optimization, and small but biologically important deviations in low-concentration nutrients would have little influence on the solution.

To avoid this imbalance, residuals are scaled relative to target magnitude. For each nutrient $i$, define a scaling factor $$
s_i =
\begin{cases}
b_i, & b_i \ne 0,\\
1, & b_i = 0.
\end{cases}
$$

For nutrients with nonzero targets, deviations are therefore measured relative to the target value. For nutrients with zero targets, deviations are measured in absolute concentration units.

To apply this scaling, each nutrient equation is divided by its corresponding factor $s_i$. In matrix form, the target-scaled design matrix is $$
A_{\mathrm{sub}} = \operatorname{diag}(1/s)\, A,
$$ and the scaled target vector is $b/s$, where division is performed elementwise.

The NNLS problem is then $$
\min_{x \ge 0}
\left\|
W \left( A_{\mathrm{sub}} x - b/s \right)
\right\|_2^2 .
$$

This is equivalent to minimizing a weighted sum of squared scaled residuals. For nutrients with nonzero targets, the optimization minimizes weighted relative deviations; for nutrients with zero targets, it minimizes weighted absolute deviations. The scaling ensures that all nutrients contribute comparably to the optimization regardless of their concentration range, while importance weights control how strictly individual targets are enforced.

Operationally, the solver predicts nutrient concentrations from candidate salt amounts, computes deviations from the targets, rescales and weights these deviations according to target magnitude and nutrient importance and determines the non-negative salt combination that minimizes the resulting weighted least-squares objective.

------------------------------------------------------------------------

### Interpretation of the solution

The NNLS solver returns the optimal non-negative salt amounts $x$. Because the optimization is performed on scaled and weighted equations, achieved nutrient concentrations are recomputed using the original, unscaled stoichiometry: $$
\hat b = A x .
$$

The vector $x$ therefore represents the physically meaningful fertilizer composition and is interpreted directly as mmol L$^{-1}$ of each salt to add to the nutrient solution. These values can therefore be used directly to prepare the nutrient solution.

------------------------------------------------------------------------

### Error quantities reported by `NutriCalc`

After computing achieved nutrient concentrations, deviations from the target composition are quantified using several complementary error measures. These metrics describe both absolute and relative discrepancies between predicted and desired nutrient levels.

The absolute error for nutrient $i$ is

$$
e_i = \hat b_i - b_i ,
$$

expressed in mmol L$^{-1}$. This measures the direct concentration difference between achieved and target values.

Percent error expresses the deviation relative to the target concentration:

$$
p_i = 100 \cdot \frac{e_i}{b_i}.
$$

Percent error is computed elementwise. Non-finite values (e.g. division by zero) are recorded as missing. If $b_i = 0$ and $|\hat b_i| < 10^{-9}$, the percent error is set to $0$.

To summarize overall deviation across all nutrients, sum of squared error is computed as

$$
\mathrm{SSE} = \sum_{i=1}^{m} e_i^2 .
$$

This metric reflects the total magnitude of absolute concentration differences.

Relative squared error evaluates deviations relative to target magnitude. Relative residuals are defined as

$$
r_i = \frac{e_i}{b_i},
$$

and non-finite values are omitted. The aggregate measure is reported as the sum of squared relative errors

$$
\mathrm{SSRE} = \sum_{i=1}^{m} r_i^2 .
$$

This metric summarizes proportional deviations across nutrients and provides a scale-independent measure of solution accuracy. Nutrients with zero target concentration where excluded from relative error metrics.

### Two-stage NNLS solver: separation of neutral salts and acid–base reagents

In addition to the standard optimization, `NutriCalc` provides an optional two-stage solver that incorporates acid–base reagents. These compounds are included in the global stoichiometry matrix and distinguished from neutral fertilizer salts using the classification `salt_info$is_acid_base`. The purpose of this separation is to ensure that nutrient targets are satisfied using neutral salts first and whenever possible, while acids and bases are used only to correct remaining deficits that neutral salts cannot resolve. This prevents acid–base reagents from acting as unconstrained degrees of freedom in the primary optimization and yields chemically more interpretable fertilizer compositions.

The two-stage procedure is activated only when the acid/base option is enabled. When this option is disabled, acid/base compounds are excluded from the selected compound set and nutrient optimization is performed using neutral salts only. Boric acid belongs to the neutral group in this classification, as it is treated as a micronutrient source rather than a pH-active reagent.

In the first stage, NNLS optimization is performed using only compounds classified as neutral salts. This stage uses the same weighted and target-scaled objective as the standard solver and produces achieved nutrient concentrations

$$
\hat b^{(1)}.
$$

Residual nutrient deviations are then computed as

$$
r = b - \hat b^{(1)}.
$$

Residual entries are filtered before any second-stage correction. Deviations within the relative tolerance are treated as negligible and set to zero. Residuals are also set to zero for nutrients that cannot be supplied by any acid or base (non-fixable nutrients). Negative residuals (oversupply) are set to zero so that only undersupplied nutrients remain eligible for correction.

If the two-stage mode is enabled and at least one fixable nutrient exceeds the tolerance criterion defined by

$$
r_i > \max(\texttt{tol\_abs}, \texttt{tol\_rel}\, b_i),
$$

a second NNLS optimization is performed using only acid/base compounds. The target for this stage is the filtered residual vector from stage 1. Scaling and importance weighting are applied using the same rules as in the standard solver, but computed from the residual target rather than the original nutrient targets. Stage 2 therefore supplies only remaining nutrient deficits that neutral salts could not provide.

Compound amounts from stage 1 and stage 2 are then inserted into their original compound indices to form the complete fertilizer vector (x). Final nutrient concentrations are recomputed from the full unscaled stoichiometry matrix,

$$
\hat b = A x,
$$

and all error metrics are evaluated from these final concentrations.

Accordingly, the two-stage solver acts as a conditional refinement: neutral salts determine the primary nutrient composition, and acid–base reagents are introduced only when residual deficits exceeding the tolerance criterion remain after the neutral fit. Lowering the importance of a specific nutrient reduces the penalty for deviating from its target during stage 1, allowing a residual deficit to remain primarily in that nutrient while the remaining nutrients are matched more closely. Because stage-2 optimization operates only on residual undersupply, this deficit can then be supplied during the second stage by the corresponding acid–base compounds.

## Chemical Modeling

### Water quality and input handling

Nutrient formulation was implemented using a target-based optimization workflow in NutriCalc. Users specify desired final nutrient concentrations as a nutrient target vector. The solver then computes the fertilizer salt additions needed to match this target. This approach is exact only when the initial water matrix is essentially deionized or demineralized water. In practical cultivation systems, however, source water already contains dissolved ions that contribute to the final solution and must therefore be accounted for before fertilizer optimization.

To represent this baseline contribution, water chemistry is modeled as a water vector containing pre-existing ion concentrations. The formulation step uses an adjusted target vector obtained by subtracting the baseline water vector from the desired nutrient target vector. For each nutrient ion $i$, the solver input is defined as:

$$
T_i^{adj} = \max\!\left(T_i^{target} - W_i,\; 0\right)
$$

where $T_i^{target}$ is the desired final concentration and $W_i$ is the concentration already present in the source water. Negative adjusted concentrations are not physically meaningful because fertilizer salts can only be added. Therefore, adjusted targets are bounded below by zero. The lower bound at zero reflects the additive-only fertilizer model used in the optimizer; excess baseline ions cannot be removed within this step. The resulting adjusted vector is then passed to the salt optimizer.

In addition to ionic concentrations, two buffering descriptors are important for water characterization: acid capacity (Säurekapazität, KS 4.3) and base capacity (Basenkapazität, KB 8.2). These are not treated as primary fertilizer targets, but are retained as water properties because they influence downstream interpretation of pH behavior.

Because users may operate with blended feed waters, NutriCalc supports mixing of multiple source waters prior to target subtraction. Each source is represented as a separate composition vector and the composite baseline is computed by mass balance. For two source waters with compositions $W_1$ and $W_2$ and volumes $V_1$ and $V_2$, the mixed baseline composition is:

$$
W^{mix} = \frac{V_1 W_1 + V_2 W_2}{V_1 + V_2}.
$$

This expression is evaluated component-wise for all dissolved constituents of the water vector. The adjusted nutrient target is subsequently computed as:

$$
T^{adj} = \max\!\left(T^{target} - W^{mix},\; 0\right).
$$

This enables realistic handling of scenarios such as blending municipal water with deionized water, rainwater, or other process waters before nutrient dosing.

Water quality inputs were derived from publicly available authority datasets where possible. For Berlin, NutriCalc retrieves supplier data and maps it to the internal water-vector representation. In practice, users can query location-specific baseline water data by postal code and optionally modify values to reflect local measurements or temporal variability.

Overall, this method ensures that fertilizer optimization is performed against the true chemical starting point rather than an idealized zero-background assumption. By explicitly modeling baseline composition, optional source-water mixing and buffering-related capacities, the resulting formulations are more representative of real cultivation solutions and suitable for subsequent pH and EC evaluation.

### Estimation of pH via charge balance

Solution pH is calculated using `ph_from_achieved()` based on the achieved nutrient composition returned by the solver (mmol L$^{-1}$). The calculation is performed as a charge-balance problem in which pH is treated as an unknown variable. The hydrogen ion concentration is determined such that the aqueous solution satisfies electroneutrality, expressed in milliequivalents per litre (meq L$^{-1}$), such that total positive and negative charge are equal [@Appelo.2005].

The calculation is formulated as a root-finding problem in which the charge-balance residual is expressed as a function of pH. The root variable is defined as

$$
\mathrm{pH}_c = -\log_{10}([\mathrm{H}^+]).
$$

For each trial value of $\mathrm{pH}_c$, equilibrium speciation is computed and the resulting charge imbalance is evaluated. The solution pH corresponds to the value of $\mathrm{pH}_c$ for which the charge residual equals zero within numerical tolerance.

Strong ions (e.g., $\mathrm{K}^+$, $\mathrm{Na}^+$, $\mathrm{Ca}^{2+}$, $\mathrm{Mg}^{2+}$, $\mathrm{NO}_3^-$, $\mathrm{Cl}^-$) contribute with fixed concentrations from the achieved nutrient composition, whereas variable ions (e.g., $\mathrm{H}^+$, $\mathrm{OH}^-$, $\mathrm{NH}_4^+$) are determined from equilibrium speciation. Weak acid–base systems (e.g., sulfate, phosphate, carbonate, borate, silicate) are included in the speciation model and contribute to the charge balance according to their equilibrium distributions.

Carbonate speciation is parameterized either from a carbonate-system input treated as total inorganic carbon when provided (e.g., `KS4_3`, interpreted approximately as $\mathrm{HCO}_3^-$), or from fixed dissolved $\mathrm{CO}_2(\mathrm{aq})$ when provided (e.g., `KB8_2`, interpreted approximately as $\mathrm{CO}_2$). Optional chelating ligands (e.g., for Fe) are included as fixed-charge anions when ligand totals are present (default net charges: EDTA, EDDHA, HBED $z=-4$ and DTPA $z=-5$).

Acid–base equilibria are evaluated using thermodynamic dissociation constants at 25 °C [@Haynes.2017]. The speciation model includes the following equilibrium systems:

| System | Equilibrium species |
|------------------------------------|------------------------------------|
| Water | $\mathrm{H}^+$, $\mathrm{OH}^-$ |
| Ammonia | $\mathrm{NH}_4^+$, $\mathrm{NH}_3$ |
| Sulfate | $\mathrm{HSO}_4^-$, $\mathrm{SO}_4^{2-}$ |
| Phosphate | $\mathrm{H}_3\mathrm{PO}_4$, $\mathrm{H}_2\mathrm{PO}_4^-$, $\mathrm{HPO}_4^{2-}$, $\mathrm{PO}_4^{3-}$ |
| Carbonate | $\mathrm{CO}_2(\mathrm{aq})$, $\mathrm{HCO}_3^-$, $\mathrm{CO}_3^{2-}$ |
| Borate | $\mathrm{B(OH)}_3$, $\mathrm{B(OH)}_4^-$ |
| Silicate | $\mathrm{H}_4\mathrm{SiO}_4$, $\mathrm{H}_3\mathrm{SiO}_4^-$ |

Non-ideal solution behavior is represented using Davies activity coefficients at 25 °C, using the Debye–Hückel parameter for water ($A = 0.5085$). Ionic strength is computed as

$$
I = \tfrac{1}{2} \sum c_i z_i^2.
$$

For each trial $\mathrm{pH}_c$, ionic strength and activity coefficients are determined iteratively because they depend on the speciation they influence. Ionic strength is computed from species concentrations, activity coefficients are updated from ionic strength and speciation is recalculated. This fixed-point iteration is repeated until successive estimates converge. The resulting activity coefficients are then applied to all equilibrium expressions.

After convergence of the inner speciation loop, the charge residual is evaluated as total positive equivalents minus total negative equivalents. Positive charge includes fixed cations and variable $\mathrm{H}^+$ and $\mathrm{NH}_4^+$. Negative charge includes fixed anions, $\mathrm{OH}^-$ and the equivalent contributions of sulfate, phosphate, carbonate, borate and silicate species, together with optional fixed chelate charge.

The bisection search is initialized over a bounded pH interval (default 3–9), reflecting the typical range of horticultural nutrient solutions and improving numerical stability. If the charge-balance function is not bracketed within this interval, the search range is expanded automatically.

The reported pH is activity-based,

$$
\mathrm{pH} = -\log_{10}\!\bigl(\gamma_H [\mathrm{H}^+]\bigr),
$$

and therefore may differ slightly from $\mathrm{pH}_c$. All calculations assume 25 °C and exclude precipitation reactions and explicit metal–ligand complexation beyond optional fixed-charge chelate terms.

### EC estimation using ionic diffusion coefficients

Electrical conductivity (EC) quantifies the ability of an aqueous solution to transport electrical charge and is therefore determined by the concentration and mobility of dissolved ions. Several approaches exist for estimating EC from chemical composition, ranging from simple empirical relationships to ion-specific physicochemical models.

In practical greenhouse fertilization, EC is often estimated from the total ionic charge concentration expressed in milliequivalents per liter (meq L$^{-1}$). A widely used rule of thumb for greenhouse nutrient solutions is [@vanderLugt.2020]:

$$
\mathrm{EC}\;(\mathrm{mS\,cm^{-1}})
=
\frac{
\sum \mathrm{meq\;cations}
+
\sum \mathrm{meq\;anions}
}{20}.
$$

Because aqueous solutions satisfy electroneutrality,

$$
\sum \mathrm{meq\;cations}
\approx
\sum \mathrm{meq\;anions},
$$

this relationship can also be interpreted as a proportionality between EC and total ionic charge concentration. A closely related approximation for dilute aqueous solutions is described by @Appelo.2005, where conductivity in µS cm$^{-1}$ is approximately proportional to total milliequivalents per liter (typically valid below about 1.5 mS cm$^{-1}$):

$$
\sum \mathrm{meq\;ions}
\;\approx\;
\frac{\mathrm{EC}\;(\mu\mathrm{S\,cm^{-1}})}{100}.
$$

More refined empirical relationships have been proposed for greenhouse nutrient solutions with relatively constrained ionic composition. Based on experimental calibration, @Sonneveld.1994 proposed the following regression relating EC to total cation concentration:

$$
\mathrm{EC}
=
0.095
\cdot
\sum \mathrm{meq\;cations}
+
0.19,
$$

where EC is expressed in mS cm$^{-1}$.

Although such empirical formulations may improve predictive accuracy within their calibration range, they do not explicitly account for differences in ionic mobility among individual species. Because the ionic species present in solution are known in `NutriCalc`, EC can be estimated mechanistically from their concentration, charge and mobility. The limiting molar ionic conductivity can be related to the diffusion coefficient through the Nernst–Einstein relation [@Kalka.2020]. In `NutriCalc`, this relation is extended by a mobility correction term accounting for non-ideal solution behavior. This method is used in hydrochemical programs such as PHREEQC (up to version 3.8) and AQION 8.7.15 [@Appelo.2010; @Kalka.2020].

\begin{align}
\label{eq:ec}
\mathrm{EC}
=
\left(\frac{F^2}{R T}\right)
\sum_{i} D_i z_i^{2}\, (\gamma_i)^{\alpha_i}\, c_i.
\tag{6}
\end{align}

\noindent where

\begin{align*}
F &\text{ is Faraday’s constant } [9.6485\times 10^{4}\ \mathrm{C\,mol^{-1}}],\\
R &\text{ is the gas constant } [8.31446\ \mathrm{J\,mol^{-1}\,K^{-1}}],\\
T &\text{ is absolute temperature } [T = t_{\mathrm{C}} + 273.15\ \mathrm{K}],\\
D_i &\text{ is the diffusion coefficient of ion } i\ [\mathrm{m^2\,s^{-1}}],\\
z_i &\text{ is the ionic charge number},\\
\gamma_i &\text{ is the activity coefficient of ion } i,\\
\alpha_i &\text{ is an empirical mobility scaling exponent},\\
c_i &\text{ is the molar concentration of ion } i.
\end{align*}

Concentrations used in the conductivity summation are expressed in mol m$^{-3}$. Concentrations obtained in mmol L$^{-1}$ are numerically identical to mol m$^{-3}$ and are therefore used directly without conversion. For conductivity, concentrations are expressed in mol m$^{-3}$ whereas ionic strength for activity modeling is calculated from concentrations expressed in mol L$^{-1}$. Diffusion coefficients in the internal data tables are stored in units of $10^{-5}\,\mathrm{cm^2\,s^{-1}}$ at 25 °C and converted to SI units prior to calculation:

$$
1\times10^{-5}\,\mathrm{cm^2\,s^{-1}}
=
1\times10^{-9}\,\mathrm{m^2\,s^{-1}}.
$$

The activity of a dissolved species depends not only on its concentration but also on the ionic strength of the solution [@Amelung.2018]. Ionic strength $(I)$ is calculated for activity modeling from molar concentrations expressed in mol L$^{-1}$:

$$
I
=
\frac{1}{2}\sum_i z_i^2 c_i.
$$

Because the classical Nernst–Einstein relation applies strictly at infinite dilution, a mobility scaling factor is introduced to account empirically for inter-ionic interactions at finite ionic strength as described by @Appelo.2010. The scaling exponent is defined as

$$
\alpha_i
=
\begin{cases}
\frac{0.6}{\sqrt{|z_i|}}, & I \le 0.36|z_i| \\
\frac{\sqrt{I}}{|z_i|}, & I > 0.36|z_i|
\end{cases}
$$

providing an ionic-strength-dependent attenuation of individual ionic contributions.

Activity coefficients $(\gamma_i)$ are obtained using the selected activity model implemented in `NutriCalc`. Two models are currently implemented: the limiting Debye–Hückel model and the Davies model [@Kalka.2023]. In the limiting Debye–Hückel model, the activity coefficient is given by

$$
-\log_{10}(\gamma_i)
=
A z_i^2 \sqrt{I},
$$

whereas in the Davies model it is given by

$$
-\log_{10}(\gamma_i)
=
A z_i^2 \left(\frac{\sqrt{I}}{1+\sqrt{I}} - 0.3I\right),
$$

where $A = 0.5085$ at 25 °C, $z_i$ is the ionic charge number and $I$ is the ionic strength in mol L$^{-1}$.

The limiting Debye–Hückel formula is typically applied at ionic strengths up to 0.1 mol L$^{-1}$ whereas the Davies formula is generally considered applicable up to about 0.5 mol L$^{-1}$. Currently the Davies activity correction is used as the default in the `NutriCalc` Shiny app.

The calculation is applied to the fully speciated aqueous ion composition obtained after nutrient formulation and equilibrium pH calculation. All dissolved charged species, including hydrogen and hydroxide ions, contribute to the conductivity sum according to their diffusion properties and concentrations.

Diffusion coefficients of ionic species at 25 °C were retrieved from @Haynes.2017. The approach assumes fully dissociated strong electrolytes and does not account for neutral species or metal–ligand complexes. Consequently, the model provides a physically based estimate of conductivity that resolves species-specific contributions but may deviate from measured EC where complexation, ion pairing or strong ion association is significant.

The osmotic potential is estimated from the calculated EC value using the empirical relationship $\Psi_{\mathrm{osm}} = -36 \cdot \mathrm{EC}$, where $\Psi_{\mathrm{osm}}$ is expressed in kPa and EC in mS cm$^{-1}$. This relationship is given by @George.2012 based on @UnitedStatesSalinityLaboratoryStaff.1954, where it was originally formulated as a relation between osmotic pressure and the electrical conductivity of the saturation extract.

## Shiny Application Workflow

### Installation or launching the web application

To install and run the application locally, the package `nutricalc` can be installed from GitHub using the package `remotes` and the following code:

```{r}
#| echo: true
#| eval: false
install.packages("remotes")
remotes::install_github("diolsa/nutricalc")
library(nutricalc)
run_app()
```

Alternatively, the deployed Shiny application can be accessed online without local installation at https://nutricalc.shinyapps.io/nutricalc-app/.

Once launched, the application opens with the title `NutriCalc 🌿 Nutrient Solution Formulation in Horticultural Sciences`.

### Input parameters

The left-hand panel of the Shiny graphical user interface (GUI), titled `Nutrient Targets & Fertilizers`, comprises five tabs: `Nutrient Targets`, `Water`, `Fertilizers`, `Recipes`, and `Tissue`.

#### Nutrient Targets

Within the `Nutrient Targets` tab, each nutrient is represented by three user-editable controls: a `Target` input field, a `Water` input field, and a `Priority` selector. The `Target` field defines the desired concentration in the final nutrient solution, whereas the `Water` field specifies the concentration contributed by the source water. Input values can be displayed and entered in $\mathrm{mmol\,L^{-1}}$, $\mathrm{\mu mol\,L^{-1}}$, or $\mathrm{mg\,L^{-1}}$. In addition, alkalinity-related water parameters can be entered using the `KS 4.3` and `KB 8.2` fields. The `Priority` control is implemented as a slider ranging from −2 to +2 and specifies the relative importance of each nutrient during optimization.

At the bottom of the panel, several global controls are available for workflow management. The `Set all to 0` button resets all target and water input values to zero, whereas `Reset to defaults` restores the predefined default settings. The `Use acids/bases` option activates the secondary optimization stage after completion of the initial solver step, thereby allowing the inclusion of acids and bases for further adjustment. Finally, the `Result` button initiates the solver and opens the calculated formulation output.

#### Water

The `Water` tab is used to define and modify the source-water composition. Mean water-analysis values for Berlin postal code areas (`PLZ`) from @BWB.2026 can be retrieved by entering a valid postal code (in the range from `10115` to `14199`). Using `Load Water`, the retrieved values are transferred directly to the `Water` input fields in the `Nutrient Targets` tab. Alternatively, `Mix Water` transfers the retrieved `PLZ` values to `Water 1` within the water-mixing interface.

Within the mixing interface, two separate source waters, `Water 1` and `Water 2`, can be defined. Concentrations for both sources can be entered manually. If one of the two sources is left empty, it is assumed to be deionized water. A slider located above the two input areas is used to specify the mixing ratio between `Water 1` and `Water 2`. In this way, blends of different water sources, such as tap water, drainage water, rainwater, aquaculture effluent, or other custom nutrient-containing solutions, can be simulated. By selecting `Load Mixed Water`, the composition of the resulting mixture is transferred to the `Water` input fields in the `Nutrient Targets` tab.

#### Fertilizers

The `Fertilizers` tab is used to define the fertilizer salts, acids and bases available for optimization. Compounds are displayed in a list together with their chemical formulas and names.

At the top of the tab, a `Salt search` text field is provided to filter the list. Both chemical formulas and compound names can be entered and the visibility of the listed compounds updates dynamically according to the search term. A `Clear search` button is located to the right of the search field and resets the search input, thereby restoring the full list.

Compounds can be selected or deselected manually using the checkbox list. In addition, the `Select all visible` and `Deselect all visible` buttons allow batch selection or deselection of compounds currently shown in the filtered list. Thus, these controls operate on the subset of compounds visible under the active search term rather than on the complete compound list.

Below the list, a status line provides summary information in the form `Using x selected of y visible (total z)`, where `x` denotes the number of currently selected compounds, `y` the number of currently visible compounds and `z` the total number of available compounds.

#### Recipes

The `Recipes` tab provides a collection of crop-specific and standard nutrient solutions used in horticulture. It is divided into four categories: `Olericulture`, `Fruticulture`, `Floriculture`, and `Standard Formulations`. Most recipes in the first three categories are based on @Sonneveld.1994, whereas the fourth category contains formulations compiled from @Hewitt.1966. The compilation by @Hewitt.1966 represents one of the most comprehensive collections of hydroponic nutrient-solution formulations published in the twentieth century.

Once a `Category` is selected, the corresponding `Recipe` list is displayed. By selecting a recipe and using the `Apply recipe` button, the respective values are transferred to the `Target` input fields in the `Nutrient Targets` tab. For recipes reported together with specific fertilizer salts, selection of the recipe can also modify the set of compounds selected in the `Fertilizers` tab. In addition, recipes are stored in different concentration units, with some reported in $\mathrm{mmol\,L^{-1}}$ and others in $\mathrm{mg\,L^{-1}}$.

#### Tissue

The `Tissue` tab allows formulation of a nutrient solution based on the mass-balance approach described by @Langenfeld.2022.

Tissue nutrient concentrations can be entered on a dry-mass basis and displayed in either `%` or `ppm`. In the central part of the tab, water-use efficiency (`WUE`) is calculated as dry mass divided by transpired water, with dry mass entered in `g` and transpired water in `L`. The resulting `WUE` value is displayed below the corresponding input fields.

Based on tissue nutrient concentration and `WUE`, nutrient demand per unit of transpired water is estimated and used to derive the corresponding nutrient-solution composition for analysis. In addition to the original approach of @Langenfeld.2022, total nitrogen is partitioned into nitrate and ammonium in accordance with the principle of electroneutrality. This extension is discussed in a later section.

### Output parameters

The right-hand side of the Shiny graphical user interface (GUI) contains the output panel, which is displayed after the solver is initiated. The panel is divided into five tabs: `Result`, `pH and EC`, `Solution`, `Anions` and `Cations`.

#### Result

The `Result` tab summarizes the optimization output by comparing the calculated nutrient composition with the user-defined target composition. For each nutrient, the table reports the values specified in the `Water` vector, the `Target` vector, and the values calculated by the solver in `Achieved`. In addition, a `Final` value is reported, representing the sum of the `Water` contribution and the `Achieved` contribution. Concentrations are displayed in $\mathrm{\mu mol\,L^{-1}}$ for values less than or equal to $0.1\ \mathrm{mmol\,L^{-1}}$ and in $\mathrm{mmol\,L^{-1}}$ for values above this threshold.

Deviation from the target composition is quantified using the error metrics `Abs. Error` and `Pct. Error`, representing the absolute deviation and the relative deviation in percent of `Final` from `Target`, respectively.

To evaluate the overall goodness of fit across all nutrients, deviations are additionally aggregated as squared error terms. Absolute deviations are summarized as the sum of squared errors (`SSE`), whereas relative deviations are summarized as the sum of squared relative errors (`SSRE`).

#### pH & EC

The `pH & EC` tab summarizes the calculated pH and EC values based on the `Final` composition. Calculations are performed for `25\,^\circ\mathrm{C}` using Davies activity correction, while complex formation is not considered. On the left side of the tab, the calculated `pH`, ionic strength (`mmol\,L^{-1}`), and total charge equivalents (`meq\,L^{-1}`) are reported. By selecting `Show pH charge breakdown`, an expanded view of the pH charge balance can be displayed, showing the contributions of fixed cations, fixed anions, and variable components in `meq\,L^{-1}`.

On the right side of the tab, EC is reported in both `mS\,cm^{-1}` and `\mu S\,cm^{-1}`. From the calculated EC, osmotic potential is additionally derived and reported in `kPa`. By selecting `Show EC contributions`, the individual ionic contributions to the calculated EC can be displayed. For each ion, the table reports the concentration `c` (`mM`), charge number `z`, and the contribution to conductivity `\kappa` (`mS\,cm^{-1}`).

Above the `pH` and `EC` output sections, an `Adjust concentration` slider is provided with a range from `-100%` to `+100%`. This control scales the concentration of the reported `Final` composition and thereby allows interactive assessment of the resulting changes in pH, EC, and related derived parameters.

#### Solution

The `Solution` tab reports the compounds used to obtain the `Achieved` composition. For each salt, acid, or base, the table reports the applied amount in $\mathrm{mmol\,L^{-1}}$, the calculated molar mass in $\mathrm{g\,mol^{-1}}$, and the corresponding concentration in $\mathrm{mg\,L^{-1}}$.

Above the `Fertilizer amounts` table, input fields are provided for working volume, stock factor, and stock volume. Based on these inputs, additional columns are computed for the required amount for the selected working-solution volume (`x L`) and, when valid stock parameters are provided, for the selected stock preparation (`y L` of `z`-fold stock).

Compounds are grouped according to the BAK assignment logic into `A-BAK`, `B-BAK`, and `Micro`. Assignment is rule-based: compounds containing micronutrients are assigned to `Micro`; compounds containing Ca, but no micronutrients, are assigned to `A-BAK`; compounds containing P or S, but neither micronutrients nor Ca, are assigned to `B-BAK`. Remaining compounds (`AB`) are assigned dynamically to `A-BAK` or `B-BAK` in order to balance the overall nutrient load between both tanks.

#### Anions and Cations

The `Anions` and `Cations` tabs report the relative proportions of selected anionic or cationic nutrient components on a $\mathrm{mmol\,L^{-1}}$ basis using ternary diagrams. In the `Anions` tab, the default components are `NO3-N`, `P`, and `S`, whereas in the `Cations` tab the default components are `K`, `Ca`, and `Mg`. Alternative ions can also be selected for comparison, including `Cl` for anions and `NH4-N` or `Na` for cations. The respective proportions of `Achieved` and `Target` are displayed directly in the ternary graph, and the corresponding percentage values are reported in the legend.

## Case study selection

Three case studies were selected to validate the performance of the application across a range of use cases. For this purpose, three user types with different objectives were defined.

### Case Study I: Alternative Hoagland II nutrient solution formulation

In this scenario, a plant scientist needs to prepare a classical Hoagland nutrient solution for an experiment using deionized water. The intended formulation is the macronutrient solution of Hoagland solution No. 2 according to @Hoagland.1938, consisting of 6 mM KNO₃, 4 mM Ca(NO₃)₂, 1 mM NH₄H₂PO₄, and 2 mM MgSO₄·7H₂O. The scientist requires the solution at several concentration levels corresponding to target EC values of 0.5, 1.0, 1.5, 2.0, 2.5, and 3.0 mS cm⁻¹. However, the standard calcium nitrate salt is unavailable and only calcium ammonium nitrate (CAN; 5Ca(NO₃)₂·NH₄NO₃·10H₂O) is available as a substitute.

### Case Study II: Tomato nutrient solution using Berlin-Dahlem water with acid adjustment

In this scenario, a horticultural technician needs to prepare a nutrient solution for tomato cultivation in rockwool using technical-grade fertilizers and tap water from Berlin-Dahlem (postal code 14195) instead of deionized water. The reference formulation is the tomato recipe reported by @Sonneveld.1994. Because the tap water contributes background ions and alkalinity, the two-stage solver is applied with nitric acid permitted in the second stage for acid adjustment. Calculations are performed for the preparation of 500 L nutrient solution and for 18 L of a 50-fold concentrated stock solution separated into A-BAK, B-BAK, and micronutrients.

### Case Study III: Nutrient solution with increased sodium concentration for halophytes using mixed water sources

In this scenario, a horticultural scientist needs to formulate a nutrient solution for an experiment on sodium effects in *Salicornia europaea* using tap water from Berlin postal code 10997. Sodium treatments in the literature are frequently imposed using NaCl, but chloride has to be limited to 1.5 mmol L⁻¹. As the chloride concentration of the tap water exceeds this threshold, the tap water is mixed with demineralised water. The base recipe “Propagation vegetable plants in rockwool” reported by @Sonneveld.1994 is used as the reference formulation. Ammonium nitrate (NH₄NO₃) is excluded from the set of permissible fertilizers. All calculations are performed for the preparation of 5 L stock solution.

# Results

The following result section is divided in two parts. The first part a general evaluation of the solver for a set of crop specific recopies and the performance of pH and EC calculation. This is followed by three case studies dealing with specific solving problems and are intended to showcase different scenarions and the practical applicability off the shiny app.

## General evaluation across crop specific recipes

To evaluate solver performance, a set of published crop specific nutrient targets from @Sonneveld.1994 were used as model input. These recipes specify target nutrient concentrations for various crops in mmol L⁻¹. The solver was only applied allowing salts to each recipe (no acids or bases) to compute the optimal fertilizer composition and achieved nutrient concentrations and were compared to the specified targets. The following sections summarize the results of this evaluation, including achieved nutrient profiles, error metrics, pH and EC. In addition, because @Sonneveld.1999 and @Sonneveld.2009 used one reference formulation to demonstrate a universal nutrient solution algorithm, the NutriCalc solution for this formulation was compared in detail with the published results.

### Achieved Nutrient solution formulations

All reported recipes (list of recipes is found in @tbl-crop-recipes-long-landscape_units ) were solved and achieved results are summarized in @tbl-crop-recipes-long-landscape_result_units. Squared absolute and relative errors are reported for each recipe, quantifying the deviation between achieved and target nutrient concentrations across all nutrients. All achieved results matched the target nutrient profiles with SSE and SSRE values close to zero. Recipes including Si as a target nutrient exhibited comparatively higher relative error values. In formulations using Fe, achieved Na concentrations were non-zero although Na was not specified in the literature target vectors and Na contributed to the absolute error. When Na was added to the target vector at the achieved concentration (Na target set equal to Fe), the corresponding SSE values decreased to zero. For recipes containing Si, applying the second solver approach with reduced weighting for NO₃–N resulted in solutions with an SSRE of 0.

```{r}
#| label: tbl-crop-recipes-long-landscape_result_units
#| echo: false
#| message: false
#| warning: false
#| results: asis

library(dplyr)
library(kableExtra)

# ---- ensure objects exist ----
if (!exists("recipes") || !exists("CANONICAL_NUTRIENTS")) {
  if (requireNamespace("nutricalc", quietly = TRUE)) {
    library(nutricalc)
  } else {
    source("R/constants.R")
    source("R/recipes.R")
    source("R/utils.R")
    source("R/solver.R")
    source("R/build_matrix.R")
  }
}

nutrients <- if (exists("CANONICAL_NUTRIENTS", inherits = TRUE)) {
  CANONICAL_NUTRIENTS
} else {
  nutricalc::CANONICAL_NUTRIENTS
}

if (!exists("nutrient_matrix", inherits = TRUE)) {
  if (exists("build_nutrient_matrix", inherits = TRUE) && exists("percentage_matrix", inherits = TRUE)) {
    nutrient_matrix <- build_nutrient_matrix(percentage_matrix)
  } else {
    data("nutrient_matrix", package = "nutricalc", envir = environment())
  }
}

if (!exists("salt_info", inherits = TRUE)) {
  data("salt_info", package = "nutricalc", envir = environment())
}

optimize_fn <- if (exists("optimize_nutrients", inherits = TRUE)) {
  optimize_nutrients
} else {
  nutricalc::optimize_nutrients
}

two_stage_fn <- if (exists("two_stage_optimize_nutrients", inherits = TRUE)) {
  two_stage_optimize_nutrients
} else {
  nutricalc::two_stage_optimize_nutrients
}

importance <- setNames(rep(0, length(nutrients)), nutrients)

fmt <- function(x, digits = 6) formatC(x, format = "f", digits = digits, drop0trailing = TRUE)

clean_name <- function(x) {
  x <- gsub("^[^A-Za-z]+\\s*", "", x)
  x <- gsub("\\(Reuse Drainage Water\\)", "(RDW)", x)
  x
}

cat_order <- c("Olericulture", "Floriculture", "Fruticulture")
exclude_recipes <- c("lettuce_bht")

# micronutrients (displayed in µmol L⁻¹)
micro_nutrients <- c("Na","Fe","Mn","Zn","B","Cu","Mo","Si")

rows <- list()

for (cat_name in cat_order) {
  recs <- recipes[vapply(recipes, function(r) identical(r$category, cat_name), logical(1))]
  recs <- recs[!names(recs) %in% exclude_recipes]

  for (r in recs) {

    t <- setNames(rep(0, length(nutrients)), nutrients)
    shared <- intersect(names(r$targets), nutrients)
    t[shared] <- as.numeric(r$targets[shared])

    si_non_zero <- ("Si" %in% names(t)) && isTRUE(abs(t[["Si"]]) > 0)

    importance_solver1 <- importance
    importance_solver2 <- importance
    if (si_non_zero && "NO3_N" %in% names(importance_solver2)) {
      importance_solver2[["NO3_N"]] <- -2
    }

    nm_all <- nutrient_matrix
    is_acid_base <- salt_info$is_acid_base[match(rownames(nm_all), salt_info$salt)]
    is_acid_base[is.na(is_acid_base)] <- FALSE

    # Solver 2 path (Si recipes) can keep acids/bases; otherwise use normal salts.
    nm_salts_solver2 <- if (si_non_zero) nm_all else nm_all[!is_acid_base, , drop = FALSE]
    # Solver 1 baseline always runs as normal solver (no acids/bases).
    nm_salts_solver1 <- nm_all[!is_acid_base, , drop = FALSE]

    if (!is.null(r$salts) && !isTRUE(r$salts)) {
      keep_s2 <- intersect(r$salts, rownames(nm_salts_solver2))
      if (length(keep_s2) > 0) {
        nm_salts_solver2 <- nm_salts_solver2[keep_s2, , drop = FALSE]
      }

      keep_s1 <- intersect(r$salts, rownames(nm_salts_solver1))
      if (length(keep_s1) > 0) {
        nm_salts_solver1 <- nm_salts_solver1[keep_s1, , drop = FALSE]
      }
    }

    tgt_keys_solver2 <- intersect(names(t), colnames(nm_salts_solver2))
    tgt_keys_solver1 <- intersect(names(t), colnames(nm_salts_solver1))
    sol <- if (si_non_zero) {
      two_stage_fn(
        nutrient_matrix = nm_salts_solver2,
        target = t[tgt_keys_solver2],
        importance = importance_solver2[tgt_keys_solver2]
      )
    } else {
      optimize_fn(
        nutrient_matrix = nm_salts_solver1,
        target = t[tgt_keys_solver1],
        importance = importance_solver1[tgt_keys_solver1]
      )
    }

    sq_abs <- unname(sol$squared_error)
    sq_rel <- unname(sol$rel_squared_error)

    sq_abs_display <- fmt(sq_abs, digits = 4)
    sq_rel_display <- fmt(sq_rel, digits = 4)

    if (si_non_zero) {
      sol_stage1 <- optimize_fn(
        nutrient_matrix = nm_salts_solver1,
        target = t[tgt_keys_solver1],
        importance = importance_solver1[tgt_keys_solver1]
      )
      sq_abs_stage1 <- unname(sol_stage1$squared_error)
      sq_rel_stage1 <- unname(sol_stage1$rel_squared_error)

      sq_abs_display <- paste0(sq_abs_display, " (", fmt(sq_abs_stage1, digits = 4), ")")
      sq_rel_display <- paste0(sq_rel_display, " (", fmt(sq_rel_stage1, digits = 4), ")")
    }

    achieved_full <- setNames(rep(0, length(nutrients)), nutrients)
    achieved_full[names(sol$achieved)] <- as.numeric(sol$achieved)

    # ---------- DISPLAY TRANSFORM ----------
    display_vals <- achieved_full
    micro_present <- intersect(names(display_vals), micro_nutrients)
    display_vals[micro_present] <- display_vals[micro_present] * 1000  # convert to µmol/L
    display_vals <- round(display_vals, 2)
    # --------------------------------------

    vals <- setNames(fmt(display_vals, digits = 2), names(display_vals))

    rows[[length(rows) + 1]] <- c(
      Category = cat_name,
      Recipe = paste0(clean_name(r$name), if (si_non_zero) "*" else ""),
      as.list(vals),
      `Squared absolute error` = sq_abs_display,
      `Squared relative error` = sq_rel_display
    )
  }
}

df <- bind_rows(lapply(rows, as.data.frame, stringsAsFactors = FALSE, check.names = FALSE))

nut_cols <- intersect(nutrients, names(df))
num_mat <- suppressWarnings(sapply(df[, nut_cols, drop = FALSE], as.numeric))
keep_cols <- colSums(!is.na(num_mat) & num_mat != 0) > 0

df <- df[, c("Category","Recipe", nut_cols[keep_cols],
             "Squared absolute error","Squared relative error"), drop = FALSE]

# ---------- UNIT HEADER ----------
macro_nutrients <- c("NO3_N","NH4_N","P","K","Ca","Mg","S")

nut_cols2 <- setdiff(names(df),
                     c("Category","Recipe",
                       "Squared absolute error","Squared relative error"))

macro_cols <- intersect(macro_nutrients, nut_cols2)
micro_cols <- intersect(micro_nutrients, nut_cols2)
other_cols <- setdiff(nut_cols2, c(macro_cols, micro_cols))

lab_macro <- "mmol~L⁻¹"
lab_micro <- "µmol~L⁻¹"

units_header <- c(" " = 2)
if (length(macro_cols) > 0) units_header <- c(units_header, setNames(length(macro_cols), lab_macro))
if (length(micro_cols) > 0) units_header <- c(units_header, setNames(length(micro_cols), lab_micro))
if (length(other_cols) > 0) units_header <- c(units_header, setNames(length(other_cols), " "))
units_header <- c(units_header, setNames(2, " "))  # error columns
# ---------------------------------

pretty_names <- names(df)
pretty_names[pretty_names == "NO3_N"] <- "NO₃–N"
pretty_names[pretty_names == "NH4_N"] <- "NH₄–N"

cat("\\setlength{\\LTpre}{-10pt}\n")
cat("\\setlength{\\LTpost}{0pt}\n")
cat("\\renewcommand{\\arraystretch}{0.85}\n")

tbl <- knitr::kable(
  df,
  format = "latex",
  booktabs = TRUE,
  longtable = TRUE,
  col.names = pretty_names,
  align = c("l","l",rep("c", ncol(df)-2)),
  caption = "Achieved Nutrient Solutions with SSE and SSRE. Asterisks denote two-stage optimization; parentheses indicate standard-solver errors.",
  escape = TRUE
) |>
  add_header_above(units_header, escape = FALSE, line = TRUE) |>
  kable_styling(latex_options = "repeat_header", font_size = 7) |>
  column_spec(ncol(df) - 1, bold = TRUE) |>
  column_spec(ncol(df), bold = TRUE) |>
  collapse_rows(columns = 1, latex_hline = "major", valign = "top") |>
  landscape()

print(tbl)

cat("\\renewcommand{\\arraystretch}{1}\n")
```

## pH and EC Outputs

```{r}
#| label: tbl-crop-recipes-long-landscape_ph_ec
#| echo: false
#| message: false
#| warning: false
#| results: asis

library(dplyr)
library(kableExtra)

`%||%` <- function(a, b) if (!is.null(a)) a else b

# ---- load objects / functions ----
if (!exists("recipes") || !exists("CANONICAL_NUTRIENTS")) {
  if (requireNamespace("nutricalc", quietly = TRUE)) {
    library(nutricalc)
  } else {
    source("R/constants.R")
    source("R/recipes.R")
    source("R/utils.R")
    source("R/solver.R")
    source("R/build_matrix.R")
    if (file.exists("R/ph.R")) source("R/ph.R")
    if (file.exists("R/ec.R")) source("R/ec.R")
  }
}

nutrients <- if (exists("CANONICAL_NUTRIENTS", inherits = TRUE)) CANONICAL_NUTRIENTS else nutricalc::CANONICAL_NUTRIENTS

if (!exists("nutrient_matrix", inherits = TRUE)) {
  if (exists("build_nutrient_matrix", inherits = TRUE) && exists("percentage_matrix", inherits = TRUE)) {
    nutrient_matrix <- build_nutrient_matrix(percentage_matrix)
  } else {
    data("nutrient_matrix", package = "nutricalc", envir = environment())
  }
}

if (!exists("salt_info", inherits = TRUE)) {
  data("salt_info", package = "nutricalc", envir = environment())
}

optimize_fn <- if (exists("optimize_nutrients", inherits = TRUE)) optimize_nutrients else nutricalc::optimize_nutrients
ph_fn <- if (exists("ph_from_achieved", mode = "function")) ph_from_achieved else nutricalc::ph_from_achieved
ec_fn <- if (exists("ec_from_ph", mode = "function")) ec_from_ph else nutricalc::ec_from_ph

importance <- setNames(rep(0, length(nutrients)), nutrients)

fmt <- function(x, digits = 2) {
  if (is.null(x) || length(x) == 0 || is.na(x) || !is.finite(x)) return("")
  formatC(as.numeric(x), format = "f", digits = digits, drop0trailing = TRUE)
}

clean_name <- function(x) {
  x <- gsub("^[^A-Za-z]+\\s*", "", x)
  x <- gsub("\\(Reuse Drainage Water\\)", "(RDW)", x)
  x
}

parse_reported_ec <- function(notes) {
  if (is.null(notes) || !nzchar(notes)) return(NA_real_)
  m <- regexpr("EC\\s*\\([^\\)]*\\)\\s*:\\s*([0-9]+(?:\\.[0-9]+)?)", notes, perl = TRUE)
  if (m[1] == -1) return(NA_real_)
  hit <- regmatches(notes, m)
  x <- sub("^.*:\\s*", "", hit)
  suppressWarnings(as.numeric(x))
}

# ---- output schema ----
out_cols <- c(
  "Category",
  "Recipe",
  "pH",
  "Ionic strength (mmol L⁻¹)",
  "TCE (meq L⁻¹)",
  "EC model (mS cm⁻¹)",
  "EC Sonneveld (mS cm⁻¹)",
  "EC reported (mS cm⁻¹)",
  "Osmotic pressure (kPa)"
)

cat_order <- c("Olericulture", "Floriculture", "Fruticulture")
exclude_recipes <- c("lettuce_bht")

rows <- list()

# ---- salts-only nutrient matrix (exclude acids/bases) ----
nm_salts_all <- nutrient_matrix
is_acid_base <- salt_info$is_acid_base[match(rownames(nm_salts_all), salt_info$salt)]
is_acid_base[is.na(is_acid_base)] <- FALSE
nm_salts_all <- nm_salts_all[!is_acid_base, , drop = FALSE]

for (cat_name in cat_order) {
  recs <- recipes[vapply(recipes, function(r) identical(r$category, cat_name), logical(1))]
  recs <- recs[!names(recs) %in% exclude_recipes]

  for (r in recs) {

    # targets in mmol/L on canonical nutrient set
    t <- setNames(rep(0, length(nutrients)), nutrients)
    if (!is.null(r$targets)) {
      shared <- intersect(names(r$targets), nutrients)
      t[shared] <- as.numeric(r$targets[shared])
    }

    # recipe salt restriction
    nm_salts <- nm_salts_all
    if (!is.null(r$salts) && !isTRUE(r$salts)) {
      keep_recipe <- intersect(r$salts, rownames(nm_salts))
      if (length(keep_recipe) > 0) nm_salts <- nm_salts[keep_recipe, , drop = FALSE]
    }

    tgt_keys <- intersect(names(t), colnames(nm_salts))
    sol <- optimize_fn(
      nutrient_matrix = nm_salts,
      target = t[tgt_keys],
      importance = importance[tgt_keys]
    )

    # build final_for_ph from salt additions only (water=0)
    final_for_ph <- setNames(rep(0, length(nutrients)), nutrients)

    add <- NULL
    if (!is.null(sol$amounts) && length(sol$amounts)) {
      if (is.null(names(sol$amounts))) names(sol$amounts) <- rownames(nm_salts)
      salt_names <- intersect(names(sol$amounts), rownames(nm_salts))
      if (length(salt_names)) {
        nm_used <- nm_salts[salt_names, , drop = FALSE]
        add <- as.vector(t(nm_used) %*% sol$amounts[salt_names])
        names(add) <- colnames(nm_used)
      }
    }

    pH_val <- NA_real_; I_mmol_L <- NA_real_; TCE_meq_L <- NA_real_
    EC_mS_cm <- NA_real_; osm_kPa <- NA_real_

    if (!is.null(add) && length(add)) {
      for (nm in names(add)) {
        if (!nm %in% names(final_for_ph)) final_for_ph[nm] <- 0
        cur <- final_for_ph[nm]
        if (is.na(cur)) cur <- 0
        final_for_ph[nm] <- cur + as.numeric(add[nm])
      }

      # salts-only => force carbonate terms to 0
      if (!"KS4_3" %in% names(final_for_ph)) final_for_ph["KS4_3"] <- 0
      if (!"CO2_aq" %in% names(final_for_ph)) final_for_ph["CO2_aq"] <- 0

      ph_res <- tryCatch(ph_fn(final_for_ph, phc_bracket = c(1, 13)), error = function(e) e)

      if (!is.null(ph_res) && is.finite(ph_res$pH)) {
        pH_val <- as.numeric(ph_res$pH)
        I_mmol_L <- as.numeric(ph_res$I) * 1000

        totals_meq <- ph_res$charge_breakdown$totals_meq
        if (!is.null(totals_meq) && all(c("pos_total_meq","neg_total_meq") %in% names(totals_meq))) {
          TCE_meq_L <- as.numeric((totals_meq[["pos_total_meq"]] + totals_meq[["neg_total_meq"]]) / 2)
        }

        ec_res <- tryCatch(ec_fn(final_for_ph, ph_res), error = function(e) NULL)
        if (!is.null(ec_res) && is.finite(ec_res$EC_mS_cm)) {
          EC_mS_cm <- as.numeric(ec_res$EC_mS_cm)
          osm_kPa <- as.numeric(-0.036 * ec_res$EC_uS_cm)
        }
      }
    }

    # Sonneveld linear EC estimate from TCE (meq/L)
    EC_sonneveld <- if (is.finite(TCE_meq_L)) (TCE_meq_L * 0.095 + 0.19) else NA_real_

    ec_rep <- parse_reported_ec(r$notes %||% "")

    row <- setNames(rep("", length(out_cols)), out_cols)
    row[["Category"]] <- cat_name
    row[["Recipe"]] <- clean_name(r$name %||% "")
    row[["pH"]] <- fmt(pH_val, 2)
    row[["Ionic strength (mmol L⁻¹)"]] <- fmt(I_mmol_L, 2)
    row[["TCE (meq L⁻¹)"]] <- fmt(TCE_meq_L, 2)
    row[["EC model (mS cm⁻¹)"]] <- fmt(EC_mS_cm, 2)
    row[["EC Sonneveld (mS cm⁻¹)"]] <- fmt(EC_sonneveld, 2)
    row[["EC reported (mS cm⁻¹)"]] <- if (is.finite(ec_rep)) fmt(ec_rep, 2) else ""
    row[["Osmotic pressure (kPa)"]] <- fmt(osm_kPa, 2)

    rows[[length(rows) + 1]] <- as.data.frame(as.list(row), check.names = FALSE, stringsAsFactors = FALSE)
  }
}

df <- if (length(rows) == 0) {
  as.data.frame(setNames(replicate(length(out_cols), character(0), simplify = FALSE), out_cols),
                check.names = FALSE, stringsAsFactors = FALSE)
} else {
  bind_rows(rows)
}
for (cn in out_cols) if (!cn %in% names(df)) df[[cn]] <- ""
df <- df[, out_cols, drop = FALSE]

cat("\\setlength{\\LTpre}{-10pt}\n")
cat("\\setlength{\\LTpost}{0pt}\n")
cat("\\renewcommand{\\arraystretch}{0.85}\n")

tbl <- knitr::kable(
  df,
  format = "latex",
  booktabs = TRUE,
  longtable = TRUE,
  align = c("l","l", rep("c", length(out_cols) - 2)),
  caption = "Solver-derived pH, ionic strength, total charge equivalents (TCE), electrical conductivity (EC) and osmotic pressure for crop-specific formulations (salts-only; water = 0; KS 4.3 = 0; CO2(aq) = 0). EC Sonneveld and reported EC for comparison.",
  escape = TRUE
) |>
  kable_styling(latex_options = "repeat_header", font_size = 7) |>
  collapse_rows(columns = 1, latex_hline = "major", valign = "top") |>
  landscape()

print(tbl)

cat("\\renewcommand{\\arraystretch}{1}\n")
```

```{r}
#| label: tbl-crop-recipes-long-landscape_ph_ec_models
#| echo: false
#| message: false
#| warning: false
#| results: asis

library(dplyr)
library(kableExtra)

`%||%` <- function(a, b) if (!is.null(a)) a else b

# ----------------------------
# Load objects + functions safely
# ----------------------------
# Prefer local sources so Quarto uses the current repo implementation even when
# render working directory is not the repository root.
find_repo_root <- function(start = getwd()) {
  d <- normalizePath(start, winslash = "/", mustWork = FALSE)
  repeat {
    if (file.exists(file.path(d, "nutricalc.Rproj")) || file.exists(file.path(d, "R", "ec.R"))) return(d)
    parent <- dirname(d)
    if (identical(parent, d)) return(NA_character_)
    d <- parent
  }
}

repo_root <- find_repo_root()
source_if_exists <- function(rel) {
  cand <- c(rel, if (!is.na(repo_root)) file.path(repo_root, rel) else character(0))
  cand <- unique(cand[file.exists(cand)])
  if (length(cand)) source(cand[1], local = FALSE)
  invisible(length(cand) > 0)
}

source_if_exists("R/constants.R")
source_if_exists("R/recipes.R")
source_if_exists("R/utils.R")
source_if_exists("R/solver.R")
source_if_exists("R/build_matrix.R")
source_if_exists("R/ph.R")
source_if_exists("R/ec.R")

if ((!exists("recipes") || !exists("CANONICAL_NUTRIENTS")) &&
    requireNamespace("nutricalc", quietly = TRUE)) {
  library(nutricalc)
}

nutrients <- if (exists("CANONICAL_NUTRIENTS", inherits = TRUE)) CANONICAL_NUTRIENTS else nutricalc::CANONICAL_NUTRIENTS

if (!exists("nutrient_matrix", inherits = TRUE)) {
  if (exists("build_nutrient_matrix", inherits = TRUE) && exists("percentage_matrix", inherits = TRUE)) {
    nutrient_matrix <- build_nutrient_matrix(percentage_matrix)
  } else {
    data("nutrient_matrix", package = "nutricalc", envir = environment())
  }
}

if (!exists("salt_info", inherits = TRUE)) {
  data("salt_info", package = "nutricalc", envir = environment())
}

optimize_fn <- if (exists("optimize_nutrients", mode = "function")) optimize_nutrients else nutricalc::optimize_nutrients
ph_fn <- if (exists("ph_from_achieved", mode = "function")) ph_from_achieved else nutricalc::ph_from_achieved
ec_fn <- if (exists("ec_from_ph", mode = "function")) ec_from_ph else nutricalc::ec_from_ph

# Compatibility guards for older installed builds.
ph_supports_gamma_model <- "gamma_model" %in% names(formals(ph_fn))
ec_supports_gamma_model <- "gamma_model" %in% names(formals(ec_fn))

importance <- setNames(rep(0, length(nutrients)), nutrients)

fmt <- function(x, digits = 2) {
  if (is.null(x) || length(x) == 0 || is.na(x) || !is.finite(x)) return("")
  formatC(as.numeric(x), format = "f", digits = digits, drop0trailing = TRUE)
}

clean_name <- function(x) {
  x <- gsub("^[^A-Za-z]+\\s*", "", x)
  x <- gsub("\\(Reuse Drainage Water\\)", "(RDW)", x)
  x
}

parse_reported_ec <- function(notes) {
  if (is.null(notes) || !nzchar(notes)) return(NA_real_)
  m <- regexpr("EC\\s*\\([^\\)]*\\)\\s*:\\s*([0-9]+(?:\\.[0-9]+)?)", notes, perl = TRUE)
  if (m[1] == -1) return(NA_real_)
  hit <- regmatches(notes, m)
  x <- sub("^.*:\\s*", "", hit)
  suppressWarnings(as.numeric(x))
}

# ----------------------------
# Output schema
# ----------------------------
out_cols <- c(
  "Category",
  "Recipe",
  "pH Davies",
  "pH DH",
  "I Davies (mmol/L)",
  "I DH (mmol/L)",
  "TCE Davies (meq/L)",
  "TCE DH (meq/L)",
  "EC Davies (mS/cm)",
  "EC DH (mS/cm)",
  "EC Sonneveld (mS/cm)",
  "EC reported (mS/cm)"
)

cat_order <- c("Olericulture", "Floriculture", "Fruticulture")
exclude_recipes <- c("lettuce_bht")

# ----------------------------
# salts-only nutrient matrix (exclude acids/bases)
# ----------------------------
nm_salts_all <- nutrient_matrix
is_acid_base <- salt_info$is_acid_base[match(rownames(nm_salts_all), salt_info$salt)]
is_acid_base[is.na(is_acid_base)] <- FALSE
nm_salts_all <- nm_salts_all[!is_acid_base, , drop = FALSE]


extract_ec_mS <- function(ec_obj) {
  if (is.null(ec_obj)) return(NA_real_)
  if (is.numeric(ec_obj) && length(ec_obj) >= 1 && is.finite(ec_obj[1])) return(as.numeric(ec_obj[1]))

  pick <- function(x, keys) {
    for (k in keys) {
      if (!is.null(x[[k]]) && is.finite(as.numeric(x[[k]][1]))) return(as.numeric(x[[k]][1]))
    }
    NA_real_
  }

  if (is.list(ec_obj)) {
    v <- pick(ec_obj, c("EC_mS_cm", "ec_mS_cm", "ECmS_cm", "EC", "kappa_mS_cm"))
    if (is.finite(v)) return(v)
    v_uS <- pick(ec_obj, c("EC_uS_cm", "ec_uS_cm", "ECuS_cm"))
    if (is.finite(v_uS)) return(v_uS / 1000)
  }

  if (is.data.frame(ec_obj) && nrow(ec_obj) >= 1) {
    for (k in c("EC_mS_cm", "ec_mS_cm", "EC", "kappa_mS_cm")) {
      if (k %in% names(ec_obj)) {
        v <- suppressWarnings(as.numeric(ec_obj[[k]][1]))
        if (is.finite(v)) return(v)
      }
    }
  }

  NA_real_
}

compute_ec_robust <- function(final_for_ph, ph_obj, gm) {
  if (is.null(ph_obj)) return(NULL)

  gm_chr <- if (isTRUE(gm == 1L)) c("davies", "davies_25c") else c("dh", "debye_huckel", "debye-huckel")
  calls <- list(
    function() ec_fn(final_for_ph, ph_obj, gamma_model = gm),
    function() ec_fn(achieved = final_for_ph, ph_res = ph_obj, gamma_model = gm),
    function() ec_fn(final_for_ph, ph_obj, gamma_model = gm_chr[1]),
    function() ec_fn(achieved = final_for_ph, ph_res = ph_obj, gamma_model = gm_chr[1]),
    function() ec_fn(final_for_ph, ph_obj, gamma_model = gm_chr[2]),
    function() ec_fn(achieved = final_for_ph, ph_res = ph_obj, gamma_model = gm_chr[2])
  )

  for (f in calls) {
    out <- tryCatch(f(), error = function(e) NULL)
    if (!is.null(out)) return(out)
  }
  NULL
}

get_tce <- function(ph_res) {
  if (is.null(ph_res)) return(NA_real_)
  totals <- ph_res$charge_breakdown$totals_meq
  if (is.null(totals) || !all(c("pos_total_meq", "neg_total_meq") %in% names(totals))) return(NA_real_)
  as.numeric((totals[["pos_total_meq"]] + totals[["neg_total_meq"]]) / 2)
}

rows <- list()

# ----------------------------
# Build table
# ----------------------------
for (cat_name in cat_order) {
  recs <- recipes[vapply(recipes, function(r) identical(r$category, cat_name), logical(1))]
  recs <- recs[!names(recs) %in% exclude_recipes]

  for (r in recs) {
    t <- setNames(rep(0, length(nutrients)), nutrients)
    if (!is.null(r$targets)) {
      shared <- intersect(names(r$targets), nutrients)
      t[shared] <- as.numeric(r$targets[shared])
    }

    nm_salts <- nm_salts_all
    if (!is.null(r$salts) && !isTRUE(r$salts)) {
      keep_recipe <- intersect(r$salts, rownames(nm_salts))
      if (length(keep_recipe) > 0) nm_salts <- nm_salts[keep_recipe, , drop = FALSE]
    }

    tgt_keys <- intersect(names(t), colnames(nm_salts))
    if (!length(tgt_keys)) next

    sol <- tryCatch(
      optimize_fn(
        nutrient_matrix = nm_salts,
        target = t[tgt_keys],
        importance = importance[tgt_keys]
      ),
      error = function(e) NULL
    )
    if (is.null(sol) || is.null(sol$amounts) || !length(sol$amounts)) next

    if (is.null(names(sol$amounts))) names(sol$amounts) <- rownames(nm_salts)
    salt_names <- intersect(names(sol$amounts), rownames(nm_salts))
    if (!length(salt_names)) next

    nm_used <- nm_salts[salt_names, , drop = FALSE]
    add <- as.vector(t(nm_used) %*% sol$amounts[salt_names])
    names(add) <- colnames(nm_used)

    final_for_ph <- setNames(rep(0, length(nutrients)), nutrients)
    for (nm in names(add)) {
      cur <- final_for_ph[nm]
      if (length(cur) == 0 || is.na(cur)) cur <- 0
      final_for_ph[nm] <- cur + as.numeric(add[nm])
    }

    final_for_ph["KS4_3"] <- 0
    final_for_ph["CO2_aq"] <- 0

    ph_call <- function(gm) {
      if (ph_supports_gamma_model) {
        ph_fn(final_for_ph, phc_bracket = c(1, 13), gamma_model = gm)
      } else {
        ph_fn(final_for_ph, phc_bracket = c(1, 13))
      }
    }

    ph_dav <- tryCatch(ph_call(1L), error = function(e) NULL)
    ph_dh <- tryCatch(ph_call(2L), error = function(e) NULL)

    pH_dav <- if (!is.null(ph_dav) && is.finite(ph_dav$pH)) as.numeric(ph_dav$pH) else NA_real_
    pH_dh <- if (!is.null(ph_dh) && is.finite(ph_dh$pH)) as.numeric(ph_dh$pH) else NA_real_

    I_dav_mmol <- if (!is.null(ph_dav) && is.finite(ph_dav$I)) as.numeric(ph_dav$I) * 1000 else NA_real_
    I_dh_mmol <- if (!is.null(ph_dh) && is.finite(ph_dh$I)) as.numeric(ph_dh$I) * 1000 else NA_real_

    TCE_dav <- get_tce(ph_dav)
    TCE_dh <- get_tce(ph_dh)

    ph_dav_ec <- ph_dav
    if (!is.null(ph_dav_ec)) ph_dav_ec$gamma_ions <- NULL
    ph_dh_ec <- ph_dh
    if (!is.null(ph_dh_ec)) ph_dh_ec$gamma_ions <- NULL

    ec_dav <- compute_ec_robust(final_for_ph, ph_dav_ec, 1L)
    ec_dh <- compute_ec_robust(final_for_ph, ph_dh_ec, 2L)

    EC_dav <- extract_ec_mS(ec_dav)
    EC_dh <- extract_ec_mS(ec_dh)

    EC_sonneveld <- if (is.finite(TCE_dav)) (TCE_dav * 0.095 + 0.19) else NA_real_
    ec_rep <- parse_reported_ec(r$notes %||% "")

    row <- setNames(rep("", length(out_cols)), out_cols)
    row[["Category"]] <- cat_name
    row[["Recipe"]] <- clean_name(r$name %||% "")
    row[["pH Davies"]] <- fmt(pH_dav, 2)
    row[["pH DH"]] <- fmt(pH_dh, 2)
    row[["I Davies (mmol/L)"]] <- fmt(I_dav_mmol, 2)
    row[["I DH (mmol/L)"]] <- fmt(I_dh_mmol, 2)
    row[["TCE Davies (meq/L)"]] <- fmt(TCE_dav, 2)
    row[["TCE DH (meq/L)"]] <- fmt(TCE_dh, 2)
    row[["EC Davies (mS/cm)"]] <- fmt(EC_dav, 2)
    row[["EC DH (mS/cm)"]] <- fmt(EC_dh, 2)
    row[["EC Sonneveld (mS/cm)"]] <- fmt(EC_sonneveld, 2)
    row[["EC reported (mS/cm)"]] <- if (is.finite(ec_rep)) fmt(ec_rep, 2) else ""

    rows[[length(rows) + 1]] <- as.data.frame(as.list(row), check.names = FALSE, stringsAsFactors = FALSE)
  }
}

df <- if (length(rows) == 0) {
  as.data.frame(setNames(replicate(length(out_cols), character(0), simplify = FALSE), out_cols),
    check.names = FALSE, stringsAsFactors = FALSE
  )
} else {
  bind_rows(rows)
}

df <- df %>%
  mutate(Category = factor(Category, levels = cat_order)) %>%
  arrange(Category, Recipe)

cat("\\setlength{\\LTpre}{-10pt}\n")
cat("\\setlength{\\LTpost}{0pt}\n")
cat("\\renewcommand{\\arraystretch}{0.85}\n")

tbl <- knitr::kable(
  df,
  format = "latex",
  booktabs = TRUE,
  longtable = TRUE,
  align = c("l", "l", rep("c", length(out_cols) - 2)),
  caption = "pH and EC comparison across activity-coefficient models (salts-only; water = 0; KS 4.3 = 0; CO2(aq) = 0). EC Sonneveld is the linear estimate from TCE (meq/L).",
  escape = TRUE
) |>
  kable_styling(latex_options = "repeat_header", font_size = 7) |>
  collapse_rows(columns = 1, latex_hline = "major", valign = "top") |>
  landscape()

print(tbl)

cat("\\renewcommand{\\arraystretch}{1}\n")

```

#### Performance of pH Calculation

Both γ activity correction models (Davies and Debye–Hückel) were evaluated for pH prediction against the phosphate buffer system (50 mL of 0.1 mol L⁻¹ KH₂PO₄ + x mL of 0.1 mol L⁻¹ NaOH), which spans pH 5.8–8.0 [@Haynes.2017; @Bower.1955]. In @fig-phosphate-buffer-calibration-two-salts, predicted pH values are plotted against the reference pH for increasing NaOH additions at 25 °C. The Davies model closely matched the reference values (R^2^ = 0.998), whereas the Debye–Hückel model showed weaker agreement (R^2^ = 0.880).

Diskussion: Consequently, γ Davies was selected as the default activity correction for pH prediction in NutriCalc.

```{r}
#| dev: cairo_pdf
#| label: fig-phosphate-buffer-calibration-two-salts
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Comparison of 50 mL of 0.1 mol L⁻¹ KH₂PO₄ + x mL of 0.1 mol L⁻¹ NaOH (25 °C)."

library(tibble)
library(dplyr)
library(ggplot2)

`%||%` <- function(a, b) if (!is.null(a)) a else b

# ----------------------------
# Load functions/data robustly (works outside repo root)
# ----------------------------
find_repo_root <- function(start = getwd()) {
  d <- normalizePath(start, winslash = "/", mustWork = FALSE)
  repeat {
    if (file.exists(file.path(d, "nutricalc.Rproj")) || file.exists(file.path(d, "R", "solver.R"))) return(d)
    parent <- dirname(d)
    if (identical(parent, d)) return(NA_character_)
    d <- parent
  }
}

repo_root <- find_repo_root()
source_if_exists <- function(rel) {
  cand <- c(rel, if (!is.na(repo_root)) file.path(repo_root, rel) else character(0))
  cand <- unique(cand[file.exists(cand)])
  if (length(cand)) source(cand[1], local = FALSE)
  invisible(length(cand) > 0)
}

source_if_exists("R/constants.R")
source_if_exists("R/utils.R")
source_if_exists("R/solver.R")
source_if_exists("R/build_matrix.R")
source_if_exists("R/ph.R")

if (requireNamespace("nutricalc", quietly = TRUE)) library(nutricalc)

# Ensure required objects
if (!exists("nutrient_matrix", inherits = TRUE)) {
  if (exists("build_nutrient_matrix", inherits = TRUE) && exists("percentage_matrix", inherits = TRUE)) {
    nutrient_matrix <- build_nutrient_matrix(percentage_matrix)
  } else {
    data("nutrient_matrix", package = "nutricalc", envir = environment())
  }
}
if (!exists("salt_info", inherits = TRUE)) data("salt_info", package = "nutricalc", envir = environment())

# Functions (prefer locally sourced)
opt2_fn <- if (exists("two_stage_optimize_nutrients", mode = "function")) {
  two_stage_optimize_nutrients
} else {
  nutricalc::two_stage_optimize_nutrients
}

ph_fn <- if (exists("ph_from_achieved", mode = "function")) {
  ph_from_achieved
} else {
  nutricalc::ph_from_achieved
}
ph_supports_gamma_model <- "gamma_model" %in% names(formals(ph_fn))

extract_pH <- function(out) {
  if (is.null(out)) return(NA_real_)
  if (is.numeric(out) && length(out) == 1 && is.finite(out)) return(as.numeric(out))
  if (is.list(out) && "pH" %in% names(out) && is.numeric(out$pH) && length(out$pH) == 1 && is.finite(out$pH)) return(as.numeric(out$pH))
  NA_real_
}

# pH calculation (use default temp = 25 °C; do NOT pass temp_C)
calc_ph <- function(achieved_named, gm = NULL) {
  achieved <- achieved_named
  achieved["KS4_3"] <- 0
  achieved["CO2_aq"] <- 0

  out <- tryCatch({
    if (!is.null(gm) && ph_supports_gamma_model) {
      ph_fn(achieved, phc_bracket = c(1, 13), gamma_model = gm)
    } else {
      ph_fn(achieved, phc_bracket = c(1, 13))
    }
  }, error = function(e) NULL)

  extract_pH(out)
}

# ----------------------------
# Reference phosphate buffer table
# ----------------------------
ref <- tribble(
  ~pH_ref, ~x_mL,
  5.80,  3.6,
  5.90,  4.6,
  6.00,  5.6,
  6.10,  6.8,
  6.20,  8.1,
  6.30,  9.7,
  6.40, 11.6,
  6.50, 13.9,
  6.60, 16.4,
  6.70, 19.3,
  6.80, 22.4,
  6.90, 25.9,
  7.00, 29.1,
  7.10, 32.1,
  7.20, 34.7,
  7.30, 37.0,
  7.40, 39.1,
  7.50, 40.9,
  7.60, 42.4,
  7.70, 43.5,
  7.80, 44.5,
  7.90, 45.3,
  8.00, 46.1
)

# Convert x -> nutrient targets for the 2-salt system
C_stock_mmol_L <- 100
targets_df <- ref %>%
  mutate(
    Vtot_mL = 50 + x_mL,
    P  = C_stock_mmol_L * (50 / Vtot_mL),
    K  = C_stock_mmol_L * (50 / Vtot_mL),
    Na = C_stock_mmol_L * (x_mL / Vtot_mL)
  )

# Restrict solver to exactly two salts
required_salts <- c("KH2PO4", "NaOH")
if (!all(required_salts %in% rownames(nutrient_matrix))) {
  stop(sprintf(
    "Missing required salts in nutrient_matrix: %s",
    paste(setdiff(required_salts, rownames(nutrient_matrix)), collapse = ", ")
  ), call. = FALSE)
}
nm_two <- nutrient_matrix[required_salts, , drop = FALSE]
importance <- c(P = 0, K = 0, Na = 0)

# ----------------------------
# Solve + compute pH for both gamma models
# ----------------------------
rows <- vector("list", nrow(targets_df))
for (i in seq_len(nrow(targets_df))) {
  target_i <- c(P = targets_df$P[i], K = targets_df$K[i], Na = targets_df$Na[i])

  sol <- tryCatch(
    opt2_fn(
      nutrient_matrix = nm_two,
      target = target_i,
      importance = importance[names(target_i)]
    ),
    error = function(e) NULL
  )

  if (is.null(sol)) {
    rows[[i]] <- tibble(
      x_mL = targets_df$x_mL[i],
      pH_ref = targets_df$pH_ref[i],
      pH_davies = NA_real_,
      pH_dh = NA_real_
    )
    next
  }

  achieved <- sol$achieved_full %||% sol$achieved
  rows[[i]] <- tibble(
    x_mL = targets_df$x_mL[i],
    pH_ref = targets_df$pH_ref[i],
    pH_davies = calc_ph(achieved, gm = 1L),
    pH_dh = if (ph_supports_gamma_model) calc_ph(achieved, gm = 2L) else NA_real_
  )
}
ph_wide <- bind_rows(rows)

# ----------------------------
# R2 vs reference (ASCII-only labels to avoid PDF encoding issues)
# ----------------------------
r2 <- function(y, yhat) {
  ok <- is.finite(y) & is.finite(yhat)
  if (!any(ok)) return(NA_real_)
  y <- y[ok]; yhat <- yhat[ok]
  1 - sum((y - yhat)^2) / sum((y - mean(y))^2)
}

r2_dav <- r2(ph_wide$pH_ref, ph_wide$pH_davies)
r2_dh  <- r2(ph_wide$pH_ref, ph_wide$pH_dh)

lab_ref <- "Buffer Reference at 25 °C @Haynes"
lab_dav <- sprintf("γ Davies (25 °C), R² = %.3f", r2_dav)
lab_dh  <- sprintf("γ Debye-Hueckel (25 °C), R² = %.3f", r2_dh)

# ----------------------------
# Plot: pH vs NaOH addition (x) + legend inside plot
# ----------------------------
plot_df <- bind_rows(
  tibble(x_mL = ref$x_mL, model = lab_ref, pH = ref$pH_ref),
  ph_wide %>% transmute(x_mL, model = lab_dav, pH = pH_davies),
  ph_wide %>% transmute(x_mL, model = lab_dh,  pH = pH_dh)
)

ggplot(plot_df, aes(x = x_mL, y = pH, linetype = model)) +
  # points first (so the line is visually continuous)
  geom_point(size = 0.8) +
  # round joins/caps make a visible difference in PDF viewers
  geom_line(
    linewidth = 0.4,
    lineend = "round",
    linejoin = "round"
  ) +
  theme_classic(base_size = 11) +
  theme(
    # axis strokes
    axis.line = element_line(linewidth = 0.4, lineend = "round"),
    axis.ticks = element_line(linewidth = 0.3, lineend = "round"),

    # legend inside plot (top-left)
    legend.position = c(0.02, 0.98),
    legend.justification = c(0, 1),
    legend.background = element_rect(fill = "white", colour = "black", linewidth = 0.2),
    legend.key = element_blank(),

    panel.grid = element_blank()
  ) +
  scale_x_continuous(limits = c(0, 50), expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(limits = c(5.4, 8.2), expand = expansion(mult = c(0, 0))) +
  labs(
    x = "NaOH addition (mL)",
    y = "pH",
    linetype = NULL
  )
```

#### Performance of EC Calculation

To evaluate EC calculation performance, models were compared against reference calibration EC values for KCl solutions at 25 °C. Electrical conductivity values for 0.01 and 0.1 mol L⁻¹ KCl were taken from @Haynes.2017, which reports values based on primary conductivity standards originally published by @Pratt.2001. In @tbl-kcl-ec-comparison , reference values are compared with NutriCalc EC predictions using both γ activity-correction models (Davies and Debye–Hückel). For comparison, the empirical approach by @Sonneveld.1999 and geochemical computations with PHREEQC [@Parkhurst.2013] are also shown. PHREEQC and NutriCalc with γ Davies were very close to the values reported by @Haynes.2017 at both concentrations. NutriCalc with γ Debye–Hückel slightly underestimated EC, while the underestimation was larger for the Sonneveld empirical approach.

```{r}
#| label: tbl-kcl-ec-comparison
#| tbl-cap: "Electrical conductivity of KCl solutions at 25 °C."
#| echo: false
#| message: false
#| warning: false
#| results: asis

library(dplyr)
library(kableExtra)
library(tibble)

`%||%` <- function(a, b) if (!is.null(a)) a else b

# ----------------------------
# Load functions/data robustly (prefer repo sources)
# ----------------------------
find_repo_root <- function(start = getwd()) {
  d <- normalizePath(start, winslash = "/", mustWork = FALSE)
  repeat {
    if (file.exists(file.path(d, "nutricalc.Rproj")) || file.exists(file.path(d, "R", "ec.R"))) return(d)
    parent <- dirname(d)
    if (identical(parent, d)) return(NA_character_)
    d <- parent
  }
}

repo_root <- find_repo_root()

source_if_exists <- function(rel) {
  cand <- c(rel, if (!is.na(repo_root)) file.path(repo_root, rel) else character(0))
  cand <- unique(cand[file.exists(cand)])
  if (length(cand)) source(cand[1], local = FALSE)
  invisible(length(cand) > 0)
}

source_if_exists("R/constants.R")
source_if_exists("R/utils.R")
source_if_exists("R/ph.R")
source_if_exists("R/ec.R")

if ((!exists("CANONICAL_NUTRIENTS")) && requireNamespace("nutricalc", quietly = TRUE)) {
  library(nutricalc)
}

nutrients <- if (exists("CANONICAL_NUTRIENTS", inherits = TRUE)) CANONICAL_NUTRIENTS else nutricalc::CANONICAL_NUTRIENTS
ph_fn <- if (exists("ph_from_achieved", mode = "function")) ph_from_achieved else nutricalc::ph_from_achieved
ec_fn <- if (exists("ec_from_ph", mode = "function")) ec_from_ph else nutricalc::ec_from_ph

ph_supports_gamma_model <- "gamma_model" %in% names(formals(ph_fn))
ec_supports_gamma_model <- "gamma_model" %in% names(formals(ec_fn))

# ----------------------------
# Helpers (same spirit as your working table)
# ----------------------------
extract_ec_mS <- function(ec_obj) {
  if (is.null(ec_obj)) return(NA_real_)
  if (is.numeric(ec_obj) && length(ec_obj) >= 1 && is.finite(ec_obj[1])) return(as.numeric(ec_obj[1]))

  pick <- function(x, keys) {
    for (k in keys) {
      if (!is.null(x[[k]]) && is.finite(as.numeric(x[[k]][1]))) return(as.numeric(x[[k]][1]))
    }
    NA_real_
  }

  if (is.list(ec_obj)) {
    v <- pick(ec_obj, c("EC_mS_cm", "ec_mS_cm", "ECmS_cm", "EC", "kappa_mS_cm"))
    if (is.finite(v)) return(v)
    v_uS <- pick(ec_obj, c("EC_uS_cm", "ec_uS_cm", "ECuS_cm"))
    if (is.finite(v_uS)) return(v_uS / 1000)
  }

  if (is.data.frame(ec_obj) && nrow(ec_obj) >= 1) {
    for (k in c("EC_mS_cm", "ec_mS_cm", "EC", "kappa_mS_cm")) {
      if (k %in% names(ec_obj)) {
        v <- suppressWarnings(as.numeric(ec_obj[[k]][1]))
        if (is.finite(v)) return(v)
      }
    }
  }

  NA_real_
}

compute_ec_robust <- function(achieved, ph_obj, gm) {
  if (is.null(ph_obj)) return(NULL)

  if (!ec_supports_gamma_model) {
    calls0 <- list(
      function() ec_fn(achieved, ph_obj),
      function() ec_fn(achieved = achieved, ph_res = ph_obj)
    )
    for (f in calls0) {
      out <- tryCatch(f(), error = function(e) NULL)
      if (!is.null(out)) return(out)
    }
    return(NULL)
  }

  gm_chr <- if (isTRUE(gm == 1L)) c("davies", "davies_25c") else c("dh", "debye_huckel", "debye-huckel")

  calls <- list(
    function() ec_fn(achieved, ph_obj, gamma_model = gm),
    function() ec_fn(achieved = achieved, ph_res = ph_obj, gamma_model = gm),
    function() ec_fn(achieved, ph_obj, gamma_model = gm_chr[1]),
    function() ec_fn(achieved = achieved, ph_res = ph_obj, gamma_model = gm_chr[1]),
    function() ec_fn(achieved, ph_obj, gamma_model = gm_chr[2]),
    function() ec_fn(achieved = achieved, ph_res = ph_obj, gamma_model = gm_chr[2])
  )

  for (f in calls) {
    out <- tryCatch(f(), error = function(e) NULL)
    if (!is.null(out)) return(out)
  }
  NULL
}

ph_call <- function(achieved, gm) {
  if (ph_supports_gamma_model) ph_fn(achieved, phc_bracket = c(1, 13), gamma_model = gm)
  else ph_fn(achieved, phc_bracket = c(1, 13))
}

fmt_int <- function(x) {
  x <- suppressWarnings(as.numeric(x))
  out <- rep("", length(x))
  ok <- is.finite(x)
  out[ok] <- formatC(x[ok], format = "f", digits = 0, drop0trailing = TRUE)
  out
}

# ----------------------------
# IMPORTANT: NutriCalc expects mmol/L for nutrient concentrations.
# Convert mol/L -> mmol/L for internal calls.
# ----------------------------
make_achieved_kcl <- function(c_mol_L) {
  c_mmol_L <- c_mol_L * 1000

  a <- setNames(rep(0, length(nutrients)), nutrients)
  if ("K"  %in% names(a))  a["K"]  <- c_mmol_L
  if ("Cl" %in% names(a))  a["Cl"] <- c_mmol_L

  if ("KS4_3" %in% names(a)) a["KS4_3"] <- 0
  if ("CO2_aq" %in% names(a)) a["CO2_aq"] <- 0
  a
}

# ----------------------------
# Reference values (µS/cm at 25 °C)
# ----------------------------
ref <- tibble::tribble(
  ~c_mol_L, ~CRC,  ~Sonneveld, ~PHREEQC,
   0.01,     1408,  1140,       1409,
   0.10,    12825,  9690,      12988
)

calc_row <- function(c_mol_L) {
  achieved <- make_achieved_kcl(c_mol_L)

  ph_dav <- tryCatch(ph_call(achieved, 1L), error = function(e) NULL)
  ph_dh  <- tryCatch(ph_call(achieved, 2L), error = function(e) NULL)

  ph_dav_ec <- ph_dav; if (!is.null(ph_dav_ec)) ph_dav_ec$gamma_ions <- NULL
  ph_dh_ec  <- ph_dh;  if (!is.null(ph_dh_ec))  ph_dh_ec$gamma_ions  <- NULL

  ec_dav <- compute_ec_robust(achieved, ph_dav_ec, 1L)
  ec_dh  <- compute_ec_robust(achieved, ph_dh_ec, 2L)

  # ec in mS/cm -> report µS/cm
  tibble(
    `NutriCalc Debye–Hückel` = extract_ec_mS(ec_dh)  * 1000,
    `NutriCalc Davies`       = extract_ec_mS(ec_dav) * 1000
  )
}

nutri_vals <- ref %>%
  rowwise() %>%
  do(bind_cols(tibble(c_mol_L = .$c_mol_L), calc_row(.$c_mol_L))) %>%
  ungroup()

df <- ref %>%
  left_join(nutri_vals, by = "c_mol_L") %>%
  transmute(
    `KCl (mol L⁻¹)` = c_mol_L,
    `CRC` = CRC,
    `Sonneveld` = Sonneveld,
    `NutriCalc Debye–Hückel` = `NutriCalc Debye–Hückel`,
    `NutriCalc Davies` = `NutriCalc Davies`,
    `PHREEQC` = PHREEQC
  )

df_print <- df %>% mutate(across(-`KCl (mol L⁻¹)`, fmt_int))

tbl <- knitr::kable(
  df_print,
  format = "latex",
  booktabs = TRUE,
  align = c("c", rep("c", ncol(df_print) - 1)),
  caption = NULL,
  escape = TRUE
) |>
  add_header_above(c(" " = 1, "µS cm⁻¹" = ncol(df_print) - 1)) |>
  kable_styling(latex_options = c("hold_position"), font_size = 9)

print(tbl)
```

### Comparison with published formulation

The nutrient solution algorithm described by @Sonneveld.2009 uses the cucumber recipe with reuse of drainage water published in @Sonneveld.1994 as a demonstration of a universal nutrient-solution formulation method proposed in @Sonneveld.1999. The target macronutrient concentrations for this demonstration are 11.75 mmol L⁻¹ NO₃–N, 1.00 mmol L⁻¹ NH₄–N, 1.25 mmol L⁻¹ P, 6.50 mmol L⁻¹ K, 2.75 mmol L⁻¹ Ca, 1.00 mmol L⁻¹ Mg, and 1.00 mmol L⁻¹ S. Micronutrients are not included because @Sonneveld.2009 reports fertilizer additions only for macronutrients in this example. For a direct comparison, NutriCalc was constrained to the same set of salts reported by @Sonneveld.2009. The published Sonneveld formulation includes nitric acid. NutriCalc matched the targets exactly (SSE = 0; SSRE = 0) using the one-stage solver without requiring acid addition. The same error metrics (SSE and SSRE) were computed for both the published Sonneveld formulation and the NutriCalc solution and are shown in @tbl-sonneveld-vs-nutricalc. For the composition of nutrient salts used for the two formulations, see @tbl-fertilizer-comparison.

```{r}
#| label: tbl-sonneveld-vs-nutricalc
#| echo: false
#| results: asis

library(dplyr)
library(knitr)
library(kableExtra)
library(tibble)

df <- tribble(
  ~nutrient, ~target, ~sonneveld, ~nutricalc,
  "NO₃–N", 11.75, 14.96, 11.75, 
  "NH₄–N",  1.00,  1.25,  1.00,
  "P",      1.25,  1.50,  1.25,
  "K",      6.50,  8.88,  6.50,
  "Ca",     2.75,  2.47,  2.75,
  "Mg",     1.00,  1.07,  1.00,
  "S",      1.00,  0.87,  1.00
) %>%
  mutate(
    se_s  = (sonneveld - target)^2,
    sre_s = ((sonneveld - target) / target)^2,
    se_n  = (nutricalc - target)^2,
    sre_n = ((nutricalc - target) / target)^2
  )

SSE_s  <- sum(df$se_s);  SSRE_s <- sum(df$sre_s)
SSE_n  <- sum(df$se_n);  SSRE_n <- sum(df$sre_n)

fe <- function(x) sprintf("%.4f", x)
fc <- function(x) sprintf("%.2f", x)

tbl <- df %>%
  transmute(
    Nutrient = nutrient,
    Target   = fc(target),

    Ach_S = fc(sonneveld),
    SE_S  = fe(se_s),
    SRE_S = fe(sre_s),

    Ach_N = fc(nutricalc),
    SE_N  = fe(se_n),
    SRE_N = fe(sre_n)
  ) %>%
  bind_rows(tibble(
    Nutrient = "",
    Target   = "",
    Ach_S    = "",
    SE_S     = paste0("SSE = ",  fe(SSE_s)),
    SRE_S    = paste0("SSRE = ", fe(SSRE_s)),
    Ach_N    = "",
    SE_N     = paste0("SSE = ",  fe(SSE_n)),
    SRE_N    = paste0("SSRE = ", fe(SSRE_n))
  ))

# pretty column labels (no whitespace hacks)
colnames(tbl) <- c("Nutrient","Target","Achieved","SE","SRE","Achieved","SE","SRE")

cat("\\vspace{1.5\\baselineskip}\n")

kable(
  tbl, booktabs = TRUE,
  align = c("l","r","r","r","r","r","r","r"),
  caption = "Comparison of Sonneveld formulation and NutriCalc solution."
) %>%
  add_header_above(c(" " = 2, "Sonneveld solution" = 3, "NutriCalc solution" = 3)) %>%
  kable_styling(latex_options = "hold_position", full_width = FALSE, font_size = 9)
```

tets testtest

```{r}
#| label: tbl-fertilizer-comparison
#| echo: false
#| results: asis

library(dplyr)
library(knitr)
library(kableExtra)
library(tibble)

df <- tribble(
  ~Name,                          ~Formula,                    ~Sonneveld, ~NutriCalc,
  "Calcium nitrate–ammonium nitrate decahydrate (5:1)", "5 Ca(NO₃)₂·NH₄NO₃·10 H₂O",     0.49,       0.55,
  "Ammonium nitrate",              "NH₄NO₃",                    0.76,       0.45,
  "Monopotassium phosphate", "KH₂PO₄",                    1.50,       1.25,
  "Magnesium nitrate hexahydrate", "Mg(NO₃)₂·6 H₂O",             1.07,       1.00,
  "Potassium sulfate",             "K₂SO₄",                     0.87,       1.00,
  "Nitric acid",                   "HNO₃",                      1.00,       NA,
  "Potassium nitrate",             "KNO₃",                      5.64,       3.25
)

fmt <- function(x) sprintf("%.2f", x)

tbl <- df %>%
  mutate(
    Sonneveld = fmt(Sonneveld),
    NutriCalc = ifelse(is.na(NutriCalc), "–", fmt(NutriCalc))
  )

cat("\\vspace{1.5\\baselineskip}\n")

kable(
  tbl,
  booktabs = TRUE,
  align = c("l","l","r","r"),
  caption = "Comparison of fertilizer additions in the Sonneveld and NutriCalc formulations."
) %>%
  add_header_above(c(" " = 2, "mmol L⁻¹" = 2)) %>%
  kable_styling(
    latex_options = "hold_position",
    full_width = FALSE,
    font_size = 9
  )
```

\newpage

## Case Study Results

### Case study I — Experimental scientist: classical solution reconstruction under restricted salt availability (deionized water) under different EC.

The macronutrient composition of Hoagland solution No. 2 consists of 14 mmol L⁻¹ NO₃–N, 1 mmol L⁻¹ NH₄–N, 1 mmol L⁻¹ P, 6 mmol L⁻¹ K, 4 mmol L⁻¹ Ca, 2 mmol L⁻¹ Mg and 2 mmol L⁻¹ S. In the absence of Ca(NO₃)₂ three exemplary alternative formulations using the technical salt CAN are shown in Table \ref{tab:tbl-hoagland-comparison}.

```{r}
#| echo: false
#| results: asis

library(tibble)
library(kableExtra)

comparison_table <- tibble(
  `Salt` = c(
    "Ca(NO₃)₂",
    "5Ca(NO₃)₂·NH₄NO₃·10H₂O",
    "KNO₃",
    "MgSO₄·7H₂O",
    "Mg(NO₃)₂·6H₂O",
    "NH₄H₂PO₄",
    "KH₂PO₄",
    "K₂SO₄",
    "NH₄NO₃",
    "(NH₄)₂SO₄",
    "Number of salts"
  ),
  `Original` = c(
    "4.00",
    "—",
    "6.00",
    "2.00",
    "—",
    "1.00",
    "—",
    "—",
    "—",
    "—",
    "4"
  ),
  `Alternative 1` = c(
    "—",
    "0.80",
    "1.20",
    "—",
    "2.00",
    "—",
    "1.00",
    "1.90",
    "—",
    "0.10",
    "6"
  ),
  `Alternative 2` = c(
    "—",
    "0.80",
    "1.00",
    "—",
    "2.00",
    "—",
    "1.00",
    "2.00",
    "0.20",
    "—",
    "6"
  ),
  `Alternative 3` = c(
    "—",
    "0.80",
    "1.20",
    "—",
    "2.00",
    "0.20",
    "0.80",
    "2.00",
    "—",
    "—",
    "6"
  )
)

tbl <- kbl(
  comparison_table,
  format = "latex",
  booktabs = TRUE,
  align = c("l", "c", "c", "c", "c"),
  caption = "Comparison of the original Hoagland solution No. 2 fertilizer combination and three alternative fertilizer combinations achieving the same target macronutrient composition. Values are given in mmol L⁻¹.",
  label = "tbl-hoagland-comparison",
  escape = FALSE
) |>
  row_spec(11, bold = TRUE) |>
  kable_styling(
    latex_options = "HOLD_position",
    full_width = FALSE
  )

print(tbl)

cat("\n\\vspace{1.5\\baselineskip}\n")
```

#### Achieved composition

The target macronutrient composition of Hoagland solution No. 2 was achieved using 0.80 mmol L⁻¹ 5Ca(NO₃)₂·NH₄NO₃·10H₂O, 0.80 mmol L⁻¹ KH₂PO₄, 5.20 mmol L⁻¹ KNO₃, 2.00 mmol L⁻¹ MgSO₄·7H₂O and 0.20 mmol L⁻¹ NH₄H₂PO₄. This is shown as Alternative 3 in Table \ref{tab:tbl-hoagland-comparison}. For the different EC Designs values are shown in Table \ref{tab:tbl-hoagland-ec-series}.

```{r}
#| echo: false
#| results: asis

library(tibble)
library(kableExtra)

salt_table <- tibble(
  `Salt / parameter` = c(
    "Strength (%)",
    "5Ca(NO₃)₂·NH₄NO₃·10H₂O",
    "KH₂PO₄",
    "KNO₃",
    "MgSO₄·7H₂O",
    "NH₄H₂PO₄"
  ),
  `0.5` = c("−79%", "0.17", "0.17", "1.09", "0.42", "0.04"),
  `1.0` = c("−57%", "0.34", "0.34", "2.24", "0.86", "0.09"),
  `1.5` = c("−33%", "0.54", "0.54", "3.48", "1.34", "0.13"),
  `2.0` = c("−9%",  "0.73", "0.73", "4.73", "1.82", "0.18"),
  `2.5` = c("+15%", "0.92", "0.92", "5.98", "2.30", "0.23"),
  `3.0` = c("+40%", "1.12", "1.12", "7.28", "2.80", "0.28")
)

tbl <- kbl(
  salt_table,
  format = "latex",
  booktabs = TRUE,
  align = c("l", rep("c", 6)),
  col.names = c("Salt / parameter", "EC 0.5", "EC 1.0", "EC 1.5", "EC 2.0", "EC 2.5", "EC 3.0"),
  caption = "Amounts of fertilizer salts required to prepare modified Hoagland solution No. 2 at target EC values from 0.5 to 3.0 mS cm⁻¹. Salt amounts are given in mmol L⁻¹.",
  label = "tbl-hoagland-ec-series",
  escape = FALSE
) |>
  add_header_above(c(" " = 1, "Concentration in mmol L⁻¹ for target EC (mS cm⁻¹)" = 6)) |>
  kable_styling(
    latex_options = "HOLD_position",
    full_width = FALSE
  )

print(tbl)

cat("\n\\vspace{1.5\\baselineskip}\n")
```

#### Deviations from target

No deviation from the target composition of a Hoagland solution No. 2 was observed in the achieved solutions, resulting in an SSE and SSRE of 0.00 equally for the original formulation and for the alternative formulations with different fertilizer salts.

#### pH and EC

\newpage

Hoagland solution No. 2

### Case study II — Horticultural technician: tomato nutrient formulation using Berlin-Dahlem water with acid adjustment

#### Achieved composition

For a nutrient solution formulated for tomatoes cultivated in rockwool [@Sonneveld.1994] and prepared with tap water from postal code 14195 [@BWB.2026], the two-stage solver with nitric acid addition was used. The calculated fertilizer salts are shown in @fig-tomato-salt-composition, including the amounts required for the preparation of 500 L nutrient solution and the corresponding amounts for 18 L of a 50-fold concentrated stock solution separated into A-BAK, B-BAK and micronutrients.

![Fertilizer composition calculated in the NutriCalc Shiny app for a tomato nutrient solution in rockwool using tap water from Berlin postal code 14195. The formulation is separated into A-BAK, B-BAK, and micronutrient stock solution fractions. Fertilizer requirements are given for the preparation of 500 L nutrient solution and for 18 L of 50-fold concentrated stock solution.](Thesis_screenshots/tomato2final.png){#fig-tomato-salt-composition width="15cm"}

#### Deviations from target

The tomato recipe specifies a target composition without Na, Cl, or Si. In the calculated solution, these elements were present due to their contribution from the tap water. Although the target composition was matched with an SSRE of 0, absolute deviations remained for Na, Cl, and Si, resulting in an SSE of 7.69. The absolute errors were 2.03 mmol L⁻¹ for Na, 1.86 mmol L⁻¹ for Cl, and 0.30 mmol L⁻¹ for Si. Of the Na concentration, 0.015 mmol L⁻¹ originated from Fe-EDTA.

#### pH and EC

The tap water from postal code 14195 had a calculated pH of 7.44 and an EC of 0.86 mS cm⁻¹. Besides the ions considered in the target nutrient composition, it supplied 2.02 mmol L⁻¹ Na⁺ and 1.86 mmol L⁻¹ Cl⁻, contributing 0.09 and 0.13 mS cm⁻¹, respectively, to the total EC. The electroneutral 0.3 mmol L⁻¹ Si present in the tap water did not contribute to the overall EC.

The calculated final nutrient solution had a pH of 5.82. Ionic strength amounted to 34.84 mmol L⁻¹, while the total concentration of charge equivalents was 24.61 meq L⁻¹. The calculated EC was 2.81 mS cm⁻¹, and the osmotic potential was −101.02 kPa. In the final solution, the largest ionic contribution to EC was from NO₃⁻ at 0.89 mS cm⁻¹, followed by K⁺ at 0.58 mS cm⁻¹ and SO₄²⁻ at 0.45 mS cm⁻¹

\newpage

### Case study III — Research scientist: nutrient formulation for a novel halophyte (Salicornia) using mixed water sources

#### Achieved composition

The calculated mixed water consists of 90 % tap water from Berlin postal code 10997 and 10 % demineralised water, thereby reducing the chloride concentration of the water source to 1.5 mmol L⁻¹. Based on this water composition and the reference recipe “Propagation vegetable plants in rockwool” reported by @Sonneveld.1994, chloride is set to 1.5 mmol L⁻¹, yielding a calculated sodium concentration of 5.5 mmol L⁻¹ to maintain electroneutrality. Since ammonium nitrate (NH₄NO₃) is excluded from the permissible fertilizer set, the target composition is achieved using an alternative fertilizer combination. The final fertilizer amounts are expressed for the preparation of 5 L stock solution.

![Screenshot of the fertilizer composition calculated in the NutriCalc Shiny app for a halophyte nutrient solution prepared from a mixture of tap water from Berlin postal code 10997 and deionized water. The formulation is shown separately for A-BAK, B-BAK, and the micronutrient stock solution. Fertilizer requirements are reported for the preparation of 5 L of 100-fold concentrated stock solution.](Thesis_screenshots/case3_salts.png){width="15cm"}

#### Deviations from target

#### pH and EC

# Discussion

## Solver robustness and formulation performance

Across the evaluated formulations, the solver reproduced most published target compositions exactly or with only very small residual deviations. This indicates that the implemented fertilizer matrix is sufficiently expressive for a broad range of horticultural nutrient solutions. In particular, the crop specific targets from @Sonneveld.1994 were matched with SSE and SSRE values close to zero in most cases, showing that the combination of stoichiometric matrix construction and normalised NNLS optimisation is capable of representing many practically relevant formulation problems with high fidelity.

Where deviations remained, they were chemically interpretable rather than arbitrary numerical failures. This is important because it means that residual error in the present framework should not immediately be interpreted as an algorithmic weakness. Instead, it often reflects specific properties of the fertilizer representation or of the target vector itself.

Salts used in the solver were selected on the basis of their solubility and their reported use in hydroponics. Salts that are not water soluble but may still be relevant in soil based systems were therefore not included in the implemented salt matrix.

The small SSE values reported for the crop set of @Sonneveld.1994 are mainly caused by the use of chelated Fe sources in the salt matrix. In the implemented representation, Fe chelates introduce Na as a counter ion. Accordingly, a non zero Na contribution can occur in the achieved solution even when Na is absent from the published target vector. This should not be interpreted as a substantive mismatch in the nutritionally relevant targets, but rather as a consequence of using chemically realistic Fe fertilizers. Consistent with this interpretation, SSE drops to zero when the Na target is set equal to the Fe associated Na contribution.

A second recurring case involves recipes containing Si, commonly at around 0.75 mmol L⁻¹. In the present matrix, Si is supplied via K~2~SiO~3~, which necessarily introduces additional K and also behaves as a strongly basic reagent. Consequently, exact matching depends on whether this coupled K contribution is represented in the target vector and or corrected by acid addition. When Si is omitted, or when the corresponding K contribution is accounted for explicitly, SSRE approaches zero.

This indicates that the residual mismatch in Si containing recipes is not primarily caused by numerical instability of the optimiser, but by chemically coupled fertilizer behaviour. In other words, the difficulty lies less in the NNLS procedure itself than in the fact that silicon supply through potassium silicate cannot be treated independently of its accompanying alkalinity and counter ion contribution, because it behaves effectively as a base used in the role of a salt.

Although Si is generally considered a beneficial rather than an essential element, positive effects on yield and disease suppression have been reported [@Rengel.2023; @Voogt.2001]. Practical use of potassium silicate often requires stoichiometric acid compensation. In the formulations discussed by @Voogt.2001, this was addressed by combining silicate addition with nitric acid and with corresponding adjustment of nitrate salts.

The fact that Si containing cases improved when acid and base correction was allowed in NutriCalc therefore supports rather than undermines the chemical plausibility of the framework. Taken together, the results suggest that NutriCalc performs robustly within the tested horticultural domain and that the remaining deviations can usually be traced back to identifiable chemical causes.

## Exact solvability, electroneutrality and non uniqueness

In this context, an electroneutral target is defined as a target vector for which the sum of cation and anion charge equivalents is balanced. A non electroneutral target is a target vector for which this condition is not fulfilled. Electroneutrality is a necessary but not sufficient condition for exact solvability when only electrically neutral salts are considered. Exact solvability also depends on the selected fertilizer set.

If the target is electroneutral and the available fertilizers span the required composition, the solver may find an exact solution. In such cases, both SSE and SSRE are zero, indicating that all targets have been matched exactly in both absolute and relative terms. However, this does not imply that the solution is unique. Different fertilizer combinations may produce the same final nutrient composition [@Steiner.1961]. Some alternative solutions differ only in the hydration state of a salt, for example Ca(NO~3~)~2~ versus Ca(NO~3~)~2~·4H~2~O, whereas others may involve chemically distinct fertilizer combinations. In some cases, the number of exact solutions may be effectively infinite because different salts can compensate for one another while preserving the same final nutrient profile.

Accordingly, the NNLS solver returns one admissible non negative optimum, but it does not characterise the full set of possible exact solutions. Users can nevertheless influence the returned solution by restricting the fertilizer set through the selection and deselection of salts. In addition, small fertilizer amounts may appear in the solution if they provide even a minor reduction in the objective function. Thus, an exact or near exact solution is not necessarily the chemically simplest or most practical one.

If the target is non electroneutral, it cannot be matched exactly using only electrically neutral salts. In that case, the solver returns the non negative best fit solution that minimises the residual error. Because NutriCalc normalises residuals by target magnitude, the optimisation criterion primarily minimises the sum of squared relative errors for non zero targets and the sum of squared absolute errors for zero targets.

A useful implication of this formulation is that infeasible targets are not solved in an intuitively arithmetic sense, but according to the relative structure of the objective function. When a target cannot be matched exactly, the remaining mismatch is distributed according to the row scaling and the imposed weights. Thus, residual error should not be interpreted simply as failure. It may instead indicate that the target is not electroneutral, that the selected fertilizer set does not span the target composition, or that chemically realistic fertilizer sources introduce unavoidable side contributions.

## Error metrics and interpretation of residual deviation

Reported errors quantify the deviation between target and achieved nutrient concentrations. For each nutrient, the formulation output reports the absolute error and, when the target is non zero, the relative error. Two aggregate metrics are derived: the sum of squared errors, SSE, and the sum of squared relative errors, SSRE. The per nutrient errors indicate which nutrients deviate from the target, by how much and in which direction. The squared error metrics summarise overall solution quality.

The optimisation itself is based on a row normalised non negative least squares, NNLS, formulation. Residuals are scaled by a nutrient specific factor: for nutrients with non zero targets, deviations are normalised by the corresponding target concentration. For nutrients with zero targets, a scaling factor of one is applied. As a result, all nutrients, including those with targets set to zero, are actively included in the objective function. Any non zero achieved concentration for a zero target nutrient contributes a quadratic penalty and is therefore discouraged during optimisation.

SSE reflects the total squared deviation across all nutrients and therefore captures leakage into zero target nutrients. SSRE, in contrast, summarises squared relative deviations only for nutrients with non zero targets and serves as the primary performance indicator for how closely specified targets were matched. However, both SSE and SSRE are post hoc reporting metrics. The solver decision is determined by the normalised and weighted NNLS objective. Consequently, a formulation with slightly lower relative deviation for non zero targets may still be rejected if it introduces substantial unintended supply of zero target nutrients, because this increases the overall objective value.

Without normalisation, an objective based solely on squared absolute errors would bias the optimisation toward macronutrients, since deviations in the µmol L^-1^ range would be numerically negligible compared with deviations in the mmol L^-1^ range. Row wise normalisation therefore ensures that proportional deviations are penalised on a comparable scale, preventing magnitude driven bias and enabling balanced optimisation for both macro and micronutrients.

In this sense, SSE and SSRE should not be interpreted as direct optimisation targets, but as complementary descriptors of the achieved solution. SSE is particularly informative for identifying chemically relevant leakage into nutrients that were intended to remain absent, whereas SSRE is more informative for evaluating proportional agreement with specified non zero targets.

A useful illustration is the 1:1 salt coupling case, for example NaCl or NH~4~NO~3~. If the targets are 1 mmol L^-1^ for one ion and 2 mmol L^-1^ for the coupled counter ion, relative error minimisation yields an optimum at 1.2 mmol L^-1^ rather than the arithmetic midpoint of 1.5 mmol L^-1^. This demonstrates a non intuitive but important property of the normalised objective: the optimum is shifted toward the smaller target because deviations are scaled relative to target magnitude.

## Weighting and two stage optimisation as practical control tools

Weighting is optional, so the solver can also be used in an unweighted form, in which all nutrients contribute equally after normalisation. If weights are applied, the problem becomes a weighted non negative least squares problem, in which deviations in highly weighted nutrients contribute more strongly to the objective function than deviations in nutrients with lower weights. Lowering the weight of one nutrient shifts more of the residual error to that nutrient and allows the remaining nutrients to be matched more closely. In practical terms, this means that one nutrient can be deliberately sacrificed to improve the overall fit of the others when the target profile is difficult or impossible to achieve with the selected salts.

This feature is particularly useful for chemically constrained formulations. If one nutrient cannot be matched simultaneously with all others because of stoichiometric coupling within the available fertilizers, reducing its weight may substantially improve the fit for the remaining targets. When the second solver stage is enabled, additional adjustment with acids or bases may further improve the final composition and partially compensate for such trade offs.

It was possible to achieve an exact match between target and achieved nutrient concentrations for most of the crop set of @Sonneveld.1994 using the one stage salts only solver. This demonstrates that the implemented salt matrix is sufficiently comprehensive to represent many standard horticultural target profiles.

The two stage solver nevertheless provides an important extension when an exact solution is not achievable with neutral salts alone. In this architecture, the first stage fits the composition using neutral fertilizer salts, whereas acids and bases are introduced only in a second corrective step. This separation is chemically meaningful because it prevents acids and bases from acting as unrestricted first stage degrees of freedom that could improve the numerical fit while reducing interpretability of the resulting fertilizer composition.

The results support this design particularly for pH active or stoichiometrically awkward cases, such as silicate containing formulations. The two stage framework should therefore not be understood merely as a numerical convenience, but as a practically motivated control mechanism that balances solvability, chemical interpretability and operational realism.

## Electroneutrality and pH as coupled constraints

Electroneutrality is one of the most fundamental principles in nutrient solution formulation. In any aqueous electrolyte solution, the sum of positive charges must equal the sum of negative charges. Thus, the total charge contributed by cations is balanced by the total charge contributed by anions [@Kalka.2024]. While many dissolved compounds dissociate into well defined cations and anions and therefore contribute predictably to charge balance, some nutrients exist in different ionic species depending on pH. At the same time, pH should not be understood merely as a measure of hydrogen ion concentration, but more accurately as a measure of hydrogen ion activity. From this perspective, pH can be interpreted as a dependent variable constrained by the electroneutrality condition in aqueous solution [@Stumm.1996]. Accordingly, the activities of H~3~O^+^ and OH^-^ adjust in conjunction with pH dependent speciation systems such as phosphate, sulfate, carbonate, borate and silicate to maintain charge balance, rather than being determined solely by the stoichiometric addition and neutralisation of acid and base.

Electroneutrality ultimately determines whether a target nutrient composition is physically realisable. In the optimisation procedure, non negativity of salt addition was implemented explicitly as a mathematical constraint, whereas electroneutrality was not imposed as an additional constraint. This is because fertilizer salts contribute ions in stoichiometrically balanced combinations. Acid or base addition changes pH and ionic speciation without violating the overall charge balance of the solution. Consequently, if a target nutrient profile is itself not electroneutral, no exact solution exists and the solver can only minimise the deviation from the specified targets.

This is directly relevant for interpretation of pH within the framework. pH should not be treated as an independently selectable scalar, but as an emergent property of charge balance, acid base equilibria and activities. Integrating charge balance based pH estimation therefore extends formulation from simple nutrient matching toward chemically coherent solution evaluation.

## Role of pH and EC modelling in the workflow

A notable strength of the framework is that formulation is not treated as complete once nutrient targets have been matched. Estimated pH and EC provide a second interpretive layer linking nutrient composition to broader physicochemical behaviour.

This is relevant because two solutions with similar nominal nutrient targets can differ in buffering, ionic strength, conductivity or pH response depending on fertilizer selection, water composition and corrective additions.

The pH of a nutrient solution is a critical parameter that determines the speciation of elements across their various physicochemical forms (free ions, complexes, chelates, ion pairs, solid and gaseous phases and different oxidation levels) and is therefore critical for nutrient bioavailability (@Rijck.1998).

The implemented pH estimation module provides a practical way to evaluate the expected pH of the formulated solution based on its composition and the applied acid or base correction. This allows users to anticipate whether the resulting pH will be within a desirable range for plant growth and to make informed decisions about formulation adjustments if necessary.

Within the tested ranges, the implemented pH and EC models provided useful chemically informed approximations. The Davies activity correction showed better pH agreement than the limiting Debye Hückel model in phosphate buffer comparisons. Likewise, EC estimation based on diffusion coefficients together with Davies activity correction agreed closely with reference KCl values and with PHREEQC based comparisons, whereas simpler empirical approaches tended to underestimate conductivity at higher concentrations.

These modules do not replace full geochemical simulation. Their value lies in practical workflow integration because users can evaluate not only whether a target vector can be approximated, but also what pH and EC consequences follow from the achieved formulation.

## Relevance of the mass balance approach

"Judging from their publications and statements, it would seem that many investigators have grown crops in a solution chosen largely at random, and that, where the results have been successful, they have recommended that solution as being specific for the particular crop concerned." [@Steiner.1961]

Optimal nutrient solutions can be formulated using the principle of mass balance [@Langenfeld.2022]. In contrast to empirically derived nutrient recipes, this approach seeks to relate nutrient supply directly to plant demand. Plant nutrient demand can be approximated from tissue nutrient composition in combination with plant water consumption, thereby linking absolute nutrient uptake to water uptake. @Langenfeld.2022 distinguish between active uptake (N, P, K, Mn), intermediate uptake (Mg, S, Fe, Cu, Zn, Mo, Ni, Cl) and passive uptake (Ca, B). Passively taken up nutrients such as Ca are closely related to transpired water and are therefore well suited to a mass balance approach. Since tissue analysis represents a destructive end point measurement, non linearities in active nutrient uptake kinetics, including transitions between high affinity and low affinity transport systems, are of limited relevance in this context. With both WUE and tissue analysis, the exact temporal dynamics of uptake do not need to be resolved explicitly, as the temporal dimension is subsumed into the final nutrient solution composition. Thus, although active and passive uptake differ mechanistically and kinetically, these differences converge in the relationship between cumulative water use and the plant tissue nutrient content observed at harvest.

On this basis, crop specific target concentrations for the nutrient solution can be derived. Under ideal conditions, such a formulation would maintain the residual or recirculating solution in approximate balance over time, thereby minimising the accumulation or depletion of individual nutrients.

The present framework is particularly relevant here because a biologically plausible demand vector is not automatically a chemically realisable nutrient solution. Even if tissue composition and water use define a meaningful elemental target, this still has to be translated into an ionic target and then into a feasible fertilizer composition. NutriCalc provides a transparent way to make this translation explicit.

A remaining limitation of tissue based mass balance approaches is the translation of total plant N into a chemically feasible nutrient solution target. Specifically, total N must be partitioned into nitrate and ammonium. To address this, the present work introduces a refinement of the mass balance approach in which total N is partitioned into NO~3~–N and NH~4~–N by imposing electroneutrality as an explicit formulation constraint. This step enables the derivation of nutrient solutions that are chemically plausible and electrochemically balanced. In the example considered here, based on @Langenfeld.2022, a total N concentration of 90 mg L^-1^ yields a partitioning of 82 mg L^-1^ NO~3~–N and 8 mg L^-1^ NH~4~–N.

This does not yet constitute a full physiological model of nutrient uptake, but it is a meaningful methodological step. It bridges biological demand estimation and chemically constrained fertilizer realisation within one transparent computational framework.

## Comparison with existing formulation approaches

Existing nutrient formulation practice spans fixed recipes, manual iterative balancing, spreadsheet tools and dedicated software. Classical approaches such as the universal algorithm of @Sonneveld.1999 are chemically informed and practically valuable, but they are not necessarily expressed as explicit constrained optimisation frameworks. Spreadsheet based calculators can be useful in practice, but often remain difficult to extend reproducibly and may obscure the underlying calculation logic. Other software such as HydroBuddy offers flexible nutrient formulation functionality, but NutriCalc contributes a distinct combination of explicit stoichiometric representation, weighted row normalised NNLS fitting, charge balance oriented evaluation and modular open source implementation in R.

The main contribution of NutriCalc is therefore not merely that it calculates fertilizer quantities, but that it formalises nutrient formulation as a transparent, reproducible and modifiable constrained optimisation problem. The R implementation supports programmable workflows and integration with other analyses, whereas the Shiny interface broadens accessibility for users without programming experience.

## Limitations and scope

Despite its strengths, the framework has clear limitations. First, precipitation reactions are not modeled beyond BAK separation. A target may therefore appear feasible in dissolved ion terms while being unstable in concentrated stock solutions.

Second, chemical speciation remains simplified. Major acid base systems and fixed charge chelates are represented, but a full complexation model is not implemented yet.

Third, the activity corrections used for pH and EC estimation remain approximate. Debye Hückel and Davies models are suitable for dilute to moderately concentrated systems, but they do not capture all non ideal effects at higher ionic strengths.

Fourth, non zero residuals may reflect limitations of the selected fertilizer matrix rather than universal chemical impossibility. Exact solvability is always conditional on the implemented fertilizer set, which is not yet complete.

Fifth, the optimisation currently targets nutrient fit rather than explicit minimisation of cost, number of salts or operational complexity. Although users can partly influence the returned solution through salt selection and weighting, these objectives are not yet formalised directly.

Finally, the framework is a formulation and evaluation model rather than a crop growth model. It can represent and assess nutrient solution chemistry, but it does not predict plant response, uptake kinetics or recirculating system dynamics over time.

These limitations define the scope of the framework rather than invalidating its usefulness. NutriCalc should therefore be understood as a transparent and extensible formulation tool that captures major stoichiometric and physicochemical constraints, while leaving more advanced geochemical and biological dynamics to future development.

# Conclusions and Outlook

## Scientific contribution

This thesis developed `NutriCalc`, an open-source computational R package with a Shiny interface for the formulation, analysis and optimisation of nutrient solutions in horticulture. The work formalised nutrient-solution design as a constrained linear optimisation problem and implemented this formulation using non-negative least squares. In doing so, it linked stoichiometric conversion, target-based optimisation, water-source correction and chemically informed evaluation of pH and electrical conductivity within one transparent workflow.

The results showed that a broad range of published horticultural target solutions can be reproduced exactly or with only minor residual deviations. Where deviations remained, these were generally chemically interpretable and could often be traced to fertilizer representation, electroneutrality constraints or chemically coupled sources such as potassium silicate. Beyond reproducing established recipes, the framework also demonstrated how nutrient formulation can be extended toward mass-balance-derived targets and charge-balance-based interpretation.

Taken together, the thesis shows that nutrient-solution formulation can be moved from recipe-based and partly opaque practice toward a reproducible, optimisation-based and chemically explicit computational workflow.

## Practical relevance for horticulture

The practical relevance of the framework lies in its applicability to both research and horticultural practice. For researchers, NutriCalc provides a transparent tool for constructing and comparing nutrient treatments under explicitly defined chemical constraints. For practitioners, it offers a flexible way to formulate crop-specific nutrient solutions while accounting for local water quality, available fertilizers and physicochemical properties of the final solution.

By combining an R package with a Shiny-based graphical user interface, the framework is accessible both for programmable analytical workflows and for users without coding experience. In this sense, NutriCalc contributes not only methodologically but also operationally to digital nutrient management in horticulture.

## Limitations

The scope of the present thesis is limited in several respects. Precipitation reactions and full metal-ligand complexation were not modeled explicitly, so the framework should not be interpreted as a complete geochemical simulation environment. Likewise, the pH and EC calculations remain approximations based on selected equilibrium and activity models. In addition, exact solvability always depends on the implemented fertilizer matrix, which is necessarily incomplete.

The framework is therefore best understood as a transparent formulation and evaluation tool that captures major stoichiometric and physicochemical constraints, but not the full complexity of nutrient chemistry or crop response.

## Future extensions

Future development of NutriCalc could expand both the scientific scope of the framework and its practical applicability in horticultural research and nutrient-solution management. One important direction is the expansion of the fertilizer database. This includes adding further compounds commonly used in horticultural practice, enabling users to define custom salts directly within the application and extending the system with additional micronutrients such as Ni and Co together with appropriate fertilizer sources. Such developments would increase flexibility and allow the software to be adapted to different regional fertilizers and cultivation systems.

A second direction concerns the chemical model. In addition to the current treatment of nutrient composition, pH and EC, future versions could incorporate more detailed chemical speciation and explicit precipitation modelling. This would improve applicability particularly for the formulation of concentrated stock solutions and would further strengthen the chemical realism of the framework.

A third direction concerns the optimisation framework itself. At present, formulation is based on an NNLS approach that focuses on matching target nutrient concentrations as closely as possible. Future versions could include additional constraints such as fertilizer price, thereby allowing the formulation process to reflect not only chemical targets but also economic considerations.

A fourth direction involves the representation of nutrient targets and the expansion of built-in datasets. Because the internal formulation matrix operates on a stoichiometric basis, conversion between molar and mass-based units is straightforward. NutriCalc could therefore be extended to accept target nutrient profiles not only in mmol L⁻¹ but also in mass-based units or nutrient percentages, which would improve usability for users more familiar with commercial fertilizer labels and analytical results reported in such forms. Likewise, the dataset of target nutrient profiles could be expanded to include a wider range of crops, production systems and developmental stages. For example, stage-specific target profiles could be provided for crops such as tomato, in which nutrient demand changes substantially between vegetative growth, flowering and fruit development. A simple external input format such as CSV could facilitate maintenance and extension of such datasets and would also allow users to adapt recipes to their own conditions.

A fifth direction is the further development of reporting and user interaction. From a practical perspective, a particularly useful next step would be the integration of export functions into the Shiny application. Because this thesis was developed in Quarto, PDF and CSV exports could be implemented to generate structured reports for individual nutrient formulations. These reports could include the selected fertilizer salts, achieved nutrient concentrations, deviations from target values and predicted pH and EC together with supporting tables and visualisations. This would strengthen the value of NutriCalc as both a scientific and an applied tool.

Overall, future extensions of NutriCalc could strengthen the package in three broad areas: broader input flexibility, more advanced chemical and optimisation routines and improved reporting and user interaction. The most important future extension, however, is the integration of experimental data, particularly for the mass-balance approach, in order to derive more precisely balanced and crop-specific nutrient recipes for closed hydroponic systems.

# Summary

This thesis addresses nutrient-solution formulation in horticulture, particularly in hydroponic and other soilless systems, where solution composition must be defined precisely under chemical and practical constraints. In practice, nutrient recipes are often derived from fixed formulations or manual trial-and-error adjustment. Such approaches are difficult to reproduce consistently across studies and prone to conversion errors, especially when water quality, fertilizer composition and different unit systems must be considered simultaneously.

To address this problem, the thesis developed `NutriCalc`, an open-source computational framework in R for the formulation, evaluation and optimisation of nutrient solutions. The central methodological contribution was the development of a reproducible computational pipeline that replaces ad hoc recipe formulation with mathematical optimisation. Nutrient-solution design was formalised as a constrained linear optimisation problem and solved using non-negative least squares, allowing fertilizer combinations to be calculated transparently while ensuring physically meaningful, non-negative quantities.

The framework integrated stoichiometric conversion of fertilizer compounds, mapping to canonical nutrient targets, transformation between molar and mass-based units, and correction for ions already present in the source water. In this way, NutriCalc established a consistent workflow linking fertilizer compounds, nutrient targets and practical concentration units. Beyond nutrient matching, it also incorporated chemically informed evaluation of electroneutrality, pH and electrical conductivity, connecting optimisation results to physicochemical properties relevant for experimental interpretation and practical nutrient management. A further contribution of the thesis was the implementation of the framework in two complementary forms: a programmable R package and a Shiny-based graphical user interface. This dual design supports both reproducible analytical workflows in research and interactive use in horticultural practice.

The results showed that a broad range of published horticultural target solutions could be reconstructed exactly or with only small residual deviations. Where exact fits were not possible, these deviations were generally chemically interpretable and could often be traced to identifiable constraints such as limited fertilizer choice, stoichiometric ion coupling across salts or electroneutrality requirements. The case studies further demonstrated that the framework can be applied to more complex scenarios involving alternative fertilizer availability, local water composition and mixed water sources.

Overall, the thesis showed that nutrient-solution design can be shifted from a recipe-based and partly opaque practice to a reproducible, optimisation-based and chemically explicit workflow, providing both a methodological framework for horticultural research and a practical basis for transparent nutrient management in soilless cultivation systems.

# References

::: {#refs}
:::

\clearpage
\restorefrontmatter

```{r}
#| echo: false

qmd_file <- "Thesis.qmd"   # change to your file name
bib_file <- "citations.bib"

# read qmd text
txt <- paste(readLines(qmd_file, warn = FALSE), collapse = "\n")

# extract cited keys like @Smith2020 or -@Smith2020
matches <- regmatches(
  txt,
  gregexpr("-?@[[:alnum:]_:.+/-]+", txt, perl = TRUE)
)[[1]]

used_keys <- unique(sub("^-?@", "", matches))

# read bib file lines
bib_lines <- readLines(bib_file, warn = FALSE)

# extract BibTeX entry keys from lines like:
# @article{Smith2020,
bib_keys <- sub(
  "^@[[:alpha:]]+\\{([^,]+),.*$",
  "\\1",
  grep("^@[[:alpha:]]+\\{", bib_lines, value = TRUE)
)

# show keys in .bib that are not yet cited
unused_keys <- setdiff(bib_keys, used_keys)

unused_keys
length(unused_keys)
```

# Appendix {.unnumbered}

# Declaration of Independence {.unnumbered}
